---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# vtree2

<!-- badges: start -->
<!-- badges: end -->

Tree diagrams for categorical data, based on the 
[original vtree package](https://github.com/nbarrowman/vtree) by
Nick Barrowman.

## Gallery


```{r echo=FALSE,message=FALSE,warning=FALSE,fig.width=14,fig.height=16}
library(cowplot)
library(vtree2)

vt <- vtree_from_freqtable(Titanic, Class, Survived)

p1 <- plot(vt, legend = TRUE)
p2 <- vt |>
  add_aliases(val_alias = list(Class = c("1st" = "1<sup>st</sup>",
                                    "2nd" = "2<sup>nd</sup>",
                                    "3rd" = "3<sup>rd</sup>"))) |>
  plot(layout = "proportional", richtext = TRUE)

data(titanicNA)
p3 <- vtree(titanicNA, Class, Sex, Survived) |>
  # only retain the third class passengers
  retain(path == "Class:3rd") |>
  # change how variables are displayed
  add_aliases(val_alias = list(Class = c("1st" = "First",
                                    "2nd" = "Second",
                                    "3rd" = "Third"))) |>
  # add default labels, built from aliases
  add_labels() |>
  # change the labels for the missing values
  mutate(label = gsub("NA", "Missing", label)) |>
  # add colors
  add_palette(palettes = c("Greys", "Blues", "Purples"),
              na = "grey90") |>
  # change the color for Females who survived
  # mark() is a helper function that returns TRUE for the nodes that
  # match the given condition
  mark(path == "Class:3rd/Sex:Female/Survived:Yes") |>
  mutate(fill = ifelse(mark, "red", fill)) |>
  mutate(color = ifelse(mark, "white", color)) |>
  # plotting with legend and custom margins
  plot(legend = TRUE,
       lwidth = .45,
       margins = c(0.05, 0.05, 0.25, 0.05))

library(dplyr)
library(purrr)

iris_imgs <- list(
versicolor="images/500px-Blue_Flag,_Ottawa.jpg",
setosa="images/500px-Irissetosa1.jpg",
virginica="images/500px-Iris_virginica_2.jpg")

get_grob <- function(img) {
  tf <- system.file(img, package="vtree2")
  img <- jpeg::readJPEG(tf)
  grid::rasterGrob(img)
}

iris_grobs <- lapply(iris_imgs, get_grob)

vt <- iris |>
  mutate(Long_Petals = as.character(Petal.Length > 4)) |>
  vtree(Species, Long_Petals) |>
  add_labels() |>
  mutate(label = ifelse(leaf, label, paste0("Iris\n", node_val)))
vt <- vt |>
  mutate(grob = map(pull(vt, "node_val"), ~ iris_grobs[[.x]]))

layout <- vt |>
  add_palette(palettes = c("Greens", "Blues")) |>
  add_layout(dir="tb", show_root=FALSE, lwidth=.9, lheight=.7,
             varspace=c(Species=2,Long_Petals=1))
p4 <- layout |>
  plot(margins=c(.05, .05, .05, .2))

p1234 <- cowplot::plot_grid(p1, p2, p3, p4, ncol=2)

# first, a tree with simplified labels on the leaf nodes
cases <- cases_from_freqtable(Titanic)
vt <- vtree(cases, Sex, Survived)

library(ggplot2)

# minimalistic ggplot showing how many passengers were in each class
# of the data frame df.
clasplot <- function(df) {
  ggplot(df, aes(x = Class, fill = Class)) +
    geom_bar() +
    scale_fill_discrete(drop=FALSE) +
    scale_x_discrete(drop=FALSE) +
    ylim(0, max(table(cases$Class))) +
    theme_minimal() +
    # remove labels, legend and axes
    theme(legend.position = "none",
          axis.title = element_blank(),
          axis.ticks = element_blank())
}

# use vtree_apply to produce plots
plots <- vtree_apply(cases, vt, FUN = clasplot)
grobs <- lapply(plots, ggplot2::ggplotGrob)
grobs[!pull(vt, leaf)] <- NA # plots only for the leaf nodes

# we add layout manually, so we can control it better
# this varspace arg means: reserve 4x as much space for the Survived nodes
# as for the Sex nodes
vt <- vt |> 
  add_layout(dir="tb", show_root=FALSE, lheight=.8,
             varspace=c(Sex=1,Survived=3)) |>
  mutate(shape = ifelse(leaf, "rectangle", "roundrectangle")) |>
  mutate(grob = grobs)
p5 <- plot(vt)

plot_grid(p1234, p5, ncol=1, rel_heights = c(2, 1))
```



## Installation

You can install the development version of vtree2 like so:

``` r
pak::pak("january3/vtree2")
```

### vtree2 workflow

 1. Prepare the data with `vtree()` or `vtree_from_freqtable()`. After this
    step, the data is immutable, frequencies and counts calculated will not
    change any more.
 2. Prune the tree for visualization with `prune()` or `retain()`.
 3. Add variable and variable level aliases with `add_aliases()`.
 3. Add labels and colors with `add_labels()` and `add_palette()` or by
    directly manipulating the `label`, `color` and `fill` columns of the
    vtree object. `find_nodes()` and `prune(..., mark_only = TRUE)` can be
    used to select nodes for coloring or labeling. The `summarize_by_node()`
    function can be used to produce additional per-node summaries which can
    be used for labeling or coloring.
 4. plot the tree with `plot()`.

### Basic plots

You can construct a vtree roughly from two types of data:

 * a data frame with one row per case (cases data frame)
 * a data frame with one row per combination of factor levels, and a frequency column (frequency table)

While the former ones are common in the wild, the builtin R examples are
often the latter. The following example shows how to construct a vtree from
a frequency table and plot it with `vtree2`.

```{r example}
library(vtree2)
tdf <- cases_from_freqtable(Titanic)
vt <- vtree(tdf, Class, Sex, Survived)
vt
```

`vt` is now an object of class `vtree`, which is basically `tidygraph`'s `tbl_graph` with some extra attributes. You can plot it with `plot()`:


```{r eval=FALSE}
plot(vt)
plot(vt, layout = "proportional", legend=TRUE, show_root=FALSE)
```


```{r example_plot,fig.width=16,fig.height=15,warning=FALSE,message=FALSE,echo=FALSE}
library(cowplot)
p1 <- plot(vt)
p2 <- plot(vt, layout = "proportional", legend=TRUE, show_root=FALSE)
plot_grid(p1, p2)
```




**There is more:** trees can be oriented vertically or horizontally, 
top to bottom, bottom to top, left to right, right to left;
labels and colors can be freely customized, trees can be pruned for display.

### Vtree pruning

With the `prune()` function, you can find nodes which fullfill a certain
condition and remove them (along with their children) from the vtree:

```{r example_pruning,fig.width=16,fig.height=15}
ucb <- cases_from_freqtable(UCBAdmissions)
vt <- vtree(ucb, Dept, Gender, Admit) |>
  prune(n < 150 | freq < .15)
plot(vt)
```

**There is more:** with `find_nodes()` you can find nodes which fullfill a
certain condition. The produced mask (a simple logical vector) can be used
to select nodes for changing labels or colors. With `retain()`, you can
select the nodes you want to keep and remove other nodes.

### Labelling and colors

While colors and labels can be generated automatically, it is possible to
customize them by modifying the `label` and `fill` columns of the vtree object:


```{r example_labelling,fig.width=12,fig.height=12}
vt <- vtree(tdf, Class, Survived)
vt <- vt |> add_labels() |> # add default labels
  add_palette() |>
  mutate(label = gsub("No", "Not survived", label)) |>
  mutate(fill = ifelse(node_col == "Survived" & node_val == "No",
                       "red", fill))

plot(vt)
```

**There is more:** `add_labels()` is highly customizable and you can
produce complex labels with a simple R expression using `sprintf()` or
`glue()`.


### Summaries

The tree can be used to produce per-node summaries of any data. For
example, if you have an additional variable for your cases, you can add
summary of the values as labels to the nodes. We select here only one node
(1st Class), just to zoom in on that part of the graph. You can also select
the nodes for which you want to manipulate the label, for example with
`mark()`:


```{r example_summary,fig.width=12,fig.height=8}
vt <- vtree(tdf, Class, Sex, Survived)
                                         
sm_txt <- summarize_by_node(tdf, vt, Age) |>
  as_label()

vt |> 
  add_labels() |>
  # only for the leaf nodes. leaf is a column in the vt nodes data frame
  # which contains TRUE for leafs and FALSE for other nodes
  mutate(label = ifelse(leaf,
                        paste0(label, "\n", sm_txt),
                        label)) |>
  retain(path == "Class:3rd") |>
  plot(show_root = FALSE)
```

**There is more:** `summarize_by_node()` can be used to calculate any summary
of categorical or continuous variables as a character vector which can be
then used as labels for the nodes. It produces a data frame
with per-node summary statistics (different for categorical and continuous
variables) which can be used for further analysis.



## AI disclosure

I have not used LLMs directly in writing the code or documentation.
However, I did use an LLM based chat to discuss some of the design issues
(e.g. to compare existing packages for graph representation). Also, for
code review / bug finding.

## Documentation

More docs available [here](https://january3.github.io/vtree2/).

 * [What are vtrees?](https://january3.github.io/vtree2/articles/what_are_vtrees.html)
 * [User manual](https://january3.github.io/vtree2/articles/vtree2.html)
 * [Vtree2 for vtree users](https://january3.github.io/vtree2/articles/vtree2_for_vtree_users.html)

## TODO/PROBLEMS

 * write a manual in the main vignette [-------| ] 90% complete
 * allow adjusting grob layouts in grob nodes
 * the nodes of the diagrammer are adjusted to the size of the labels. We
   could do that, actually. I think it doesn't look so good, but OK.
 * vtree constructors take only character or factor variables: ~~build in
   checking for column type~~ maybe add functions to convert numeric to
   factor? or some automation controlled with a param?
 * add_layout should not modify shape if present
 * faster processing of cases into trees
 * why is plotting patterns so slow?
 * ~~easier formatting of labels using glue syntax~~
 * ~~summaries_vt and summaries_vt_df - that is clunky, maybe as_df option or
   smth? -> rewrote it completely~~
 * ~~better support in add_palette for 1/ variable driven palettes (i.e. one
   color for one tree level / variable of the tree, deterministic such that
   colors don't change if we prune a tree) 2/ fully customized palettes~~
 * ~~original vtree tries to keep sister NA nodes of the retained nodes. This
   behavior is reproduced by the keep_na_sisters parameter, which adds NA
   nodes to the retained nodes. However, if we target an NA node with
   is.na(), then the behavior results in the NA node being kept.~~
 * ~~sameline template for add_labels()~~
 * ~~fix the mess around legend param~~
 * ~~vtree constructors should check for presence of '/' and ':' in variable
   names~~
 * ~~vtree constructors should check that all columns are factors or characters~~
 * ~~vtree constructors should test for the presence of structural columns in
   the data. e.g. if the data has already a column called "n" or "freq", we
   will have a clash when pruning and virtual columns are created.~~
 * ~~print.vtree should show the status of added layouts / colors / etc
   etc~~
 * ~~pruning:~~
    * ~~streamline and clean up pruning / selecting nodes~~
    * ~~remove mark_only parameter, enough to have "mark()"~~ -> nah, it is
      useful for debugging
    * ~~unexport find_parents / find_children~~ -> nah, they might be useful
    * ~~remove the na.rm tag~~
 * the arrows should be attached dynamically in grob.R rather than in
   layout?
 * ~~add prefix and suffix parameters to add_labels, to make the handling
   easier~~ - nah, you can just paste() it.
 * ~~how is vtree actually handling the palettes?~~
 * ~~root node should not show percentages on default labels~~
 * ~~make sure that node_name is used for displaying variable names~~
 * ~~plotting function for patterns~~
 * ~~varspace and varsize should work with layouts other than regular~~
 * ~~add the keep_children parameter to the keep() function~~, ~~and also
   keep_na. - this is harder; because not all the NA nodes are kept, but
   only these which are on the same level as a node that is kept.
   Generally make keep behave the same as in vtree.~~
 * ~~legends~~
 * ~~should the vtree() function take cases or samples? Maybe instead of
   vtree_from_freqtable, we can just use vtree(..., .freq_col="Freq") or
   something like this.~~ nah, better make it explicit.
 * ~~maybe gridtext for the labels so we can use some basic formatting~~
 * questions to N:
   - which NA nodes are kept when vp=TRUE, only the sisters or all on the
     same level?
   - what does 
     `keep=list(eligible="Eligible",randomized="Randomized",followup="Followed up")`
     *precisely* mean? How does it differ from keep=list(followup="Followed up")?


## BUGS

 * legend titles sometimes overlap with the legend
 * ~~check the graphic summary example. if dir is tb, it looks ok (fonts
   maybe slightly too large). But turn it around and the fonts on the grob
   nodes are unreadable. Why?~~
 * ~~aliases are ignored for the legend. Maybe aliases should behave like
   palettes, i.e. also set an attribute to be used by legend?~~
 * ~~"Missing" is missing from the legend on the user guide example for
   adding aliases~~
 * ~~if some variables disappear from the plot, the legend throws an
   error~~
 * ~~prune(!Severity %in% c("Mild", "Moderate"), follow_only = TRUE) does not work~~
 * ~~na.rm handling in prune() is not consistent, at least not with vtree's behavior~~
 * ~~layout="proportional" stopped working~~
 * ~~with precomputed layout, plot doesn't know that the direction is
   vertical and the legends suck -> should the legends be computed by the
   layout function? naaaah. layout should store the attributes via an
   exported function. Should we make an S3 class? vtree_layout?~~

