---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# vtree2

<!-- badges: start -->
<!-- badges: end -->

The goal of vtree2 is to ...

## Installation

You can install the development version of vtree2 like so:

``` r
pak::pak("january3/vtree2")
```

## Quick start

### What is a vtree?

### Basic plots

You can construct a vtree roughly from two types of data:

 * a data frame with one row per case (cases data frame)
 * a data frame with one row per combination of factor levels, and a frequency column (frequency table)

While the former ones are common in the wild, the builtin R examples are
often the latter. The following example shows how to construct a vtree from
a frequency table and plot it with `vtree2`.

```{r example}
library(vtree2)
tdf <- cases_from_freqtable(Titanic)
vt <- vtree(tdf, Class, Sex, Survived)
vt
```

`vt` is now an object of class `vtree`, which is basically `tidygraph`'s `tbl_graph` with some extra attributes. You can plot it with `plot()`:


```{r example_plot,fig.width=8,fig.height=5}
plot(vt)
```

### Vtree pruning


```{r example_pruning}
ucb <- cases_from_freqtable(UCBAdmissions)
vt <- vtree(ucb, Dept, Gender, Admit) |>
  prune(n < 100 | freq < .15)
plot(vt, by_freq = TRUE)
```




### Vtree patterns


## TODO

 * plotting function for patterns
 * should the vtree function take cases or samples?
