---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# vtree2

<!-- badges: start -->
<!-- badges: end -->

Tree diagrams for categorical data, based on the 
[original vtree package](https://github.com/nbarrowman/vtree) by
Nick Barrowman.

## Installation

You can install the development version of vtree2 like so:

``` r
pak::pak("january3/vtree2")
```

## Quick start

### What is a vtree?

### Basic plots

You can construct a vtree roughly from two types of data:

 * a data frame with one row per case (cases data frame)
 * a data frame with one row per combination of factor levels, and a frequency column (frequency table)

While the former ones are common in the wild, the builtin R examples are
often the latter. The following example shows how to construct a vtree from
a frequency table and plot it with `vtree2`.

```{r example}
library(vtree2)
tdf <- cases_from_freqtable(Titanic)
vt <- vtree(tdf, Class, Sex, Survived)
vt
```

`vt` is now an object of class `vtree`, which is basically `tidygraph`'s `tbl_graph` with some extra attributes. You can plot it with `plot()`:


```{r example_plot,fig.width=16,fig.height=15,warning=FALSE,message=FALSE}
plot(vt)
```

**There is more:** trees can be oriented vertically or horizontally, 
top to bottom, bottom to top, left to right, right to left;
labels and colors can be freely customized, trees can be pruned for display.

### Vtree pruning

With the `prune()` function, you can find nodes which fullfill a certain
condition and remove them (along with their children) from the vtree:

```{r example_pruning,fig.width=16,fig.height=15}
ucb <- cases_from_freqtable(UCBAdmissions)
vt <- vtree(ucb, Dept, Gender, Admit) |>
  prune(n < 150 | freq < .15)
plot(vt, proportional = TRUE)
```

**There is more:** with `find_nodes()` you can find nodes which fullfill a
certain condition. The produced mask (a simple logical vector) can be used
to select nodes for changing labels or colors. With `keep()`, you can
select the nodes you want to keep and remove other nodes.

### Labelling and colors

While colors and labels can be generated automatically, it is possible to
customize them by modifying the `label` and `fill` columns of the vtree object:


```{r example_labelling,fig.width=8,fig.height=8}
vt <- vtree(tdf, Class, Survived)
vt <- vt |> add_labels() |> # add default labels
  add_palette() |>
  mutate(label = gsub("No", "Not survived", label)) |>
  mutate(fill = ifelse(node_col == "Survived" &
                       node_val == "No", "red", fill))

plot(vt)
```

**There is more:** `add_labels()` is highly customizable and you can
produce complex labels with a simple R expression using `sprintf()` or
`glue()`.


### Summaries

The tree can be used to produce per-node summaries of any data. For
example, if you have an additional variable for your cases, you can add
summary of the values as labels to the nodes.


```{r example_summary,fig.width=14,fig.height=30}
vt <- vtree(tdf, Class, Sex, Survived)
                                         
sm_txt <- summary_vt(tdf, vt, Age)
vt |> 
  add_labels() |>
  mutate(label = paste0(label, "\n", sm_txt)) |> plot()
```

**There is more:** `summary_vt()` can be used to calculate any summary
of categorical or continuous variables as a character vector which can be
then used as labels for the nodes. `summary_vt_df()` produces a data frame
with per-node summary statistics (different for categorical and continuous
variables) which can be used for further analysis.



### Vtree patterns

## Differences between vtree2 and the original vtree

### Motivation

The original vtree package implements most of its functionality in a single
function `vtree()` producing a plot. The function implements its own
mini-language for specifying formats of labels, variables, conditions to
select nodes and so on. This allows for simple and terse function calls.

However, this limits the flexibility of the package. It is not possible to
fully customize the labels, inspect the interim data, use the various
summaries for other purposes (e.g. to produce simple tables). `vtree2`
separates data preparation, label construction, node selection and plotting
into four separate steps. This comes at the cost of more verbose function
calls.

### Quick HOWTO for vtree users

#### Data pre-processing

Several options of the original vtree are actually for pre-processing of
the data. For example, the variable specification 'variable>value' in the
vtree mini-language is used to split a numeric variable into two levels.

I think that explicing splitting of the data before constructing the vtree
is more fittin. It is easy, and for complex cases there are many
specialized tools and packages dealing with that. For most cases, a simple
`cut()` or `ifelse()` is enough.


```{r eval=FALSE}
iris |>
  mutate(Sepal.Length = ifelse(Sepal.Length > 5.5, "long", "short")) |>
  mutate(Sepal.Width = cut(Sepal.Width, breaks = c(2, 3, 4, 5))) |>
  vtree(Sepal.Length, Sepal.Width, Species) |>
  plot()

```




#### Pruning

**Pruning.** In vtree2, you can prune the tree with the `prune()` function.
It takes a logical vector of length equal to the number of nodes in the
vtree object. You can use the variables in from the vtree node data frame
to construct the logical vector. For example, to prune all nodes with
frequency less than 0.15 or number of cases less than 150, you can do
`prune(vt, n < 150 | freq < .15)`. The result is always a vtree which you
can then plot.

```{r eval=FALSE}
data(FakeData, package="vtree")

# vtree::vtree(FakeData, "Severity Sex")
vtree(FakeData, Severity, Sex) |> plot()

#vtree::vtree(FakeData,"Severity Sex",
#      prune=list(Severity=c("Mild","Moderate")))
vtree(FakeData, Severity, Sex) |>
  prune(Severity %in% c("Mild", "Moderate")) |>
  plot()
```

**`prunebelow`**

```{r eval=FALSE}
#vtree::vtree(FakeData, "Severity Sex",
#        prunebelow=list(Severity=c("Mild","Moderate")))
vtree(FakeData, Severity, Sex) |>
  prune(Severity %in% c("Mild", "Moderate"),
        follow_only=TRUE) |>
  plot()
```

**`follow`** 
```{r eval=FALSE}
#vtree::vtree(FakeData, "Severity Sex",
#        follow=list(Severity=c("Mild","Moderate")))
# XXX doesn't work yet
vtree(FakeData, Severity, Sex) |>
  prune(!Severity %in% c("Mild", "Moderate"), follow_only = TRUE) |>
  plot()
```

**Targetted pruning.** Since vtree2 accepts any expression creating a
logical vector for pruning, specifying a targetted pruning is easy.

**Keeping.** The keep parameter in vtree is a bit peculiar, because it
silently keeps also the nodes which are NA *if* the valid percentages are
used. Also, both the *children* and the *parents* of the selected nodes are
kept.

XXX

```{r eval=FALSE}
#vtree::vtree(FakeData, "Severity Sex",
#             keep=list(Severity=c("Moderate")))
vtree(FakeData, Severity, Sex) |>
  keep(Severity == "Moderate") |>
  plot()
```

**`prunesmaller`** This is for selecting nodes based on the number of
observations. In vtree2, you can use any column from the node data frame -
including `n`, the number of observations in the node, `freq`, the
calculated proportion, `denominator` etc etc.


```{r}
#vtree::vtree(FakeData, "Severity Sex Age Category", prunesmaller = 3)
# XXX: again, vtree keeps NA nodes even if they are smaller than 3
FakeData |> mutate(Age = as.character(Age)) |>
  vtree(Severity, Sex, Age, Category) |>
  prune(n < 3, na.rm = FALSE) |>
  plot()

```


#### Formatting labels

In vtree2, there are some functions that construct automatic labels, but
for any more complex case you can use standard R functions to construct
your own labels.

By default, `add_labels()` construct the labels implicitly when you call
plot() on a vtree object. However, you can add the labels yourself either
with `add_labels()` (which is quite flexible) or by directly modifying the
`label` column in the vtree. `plot.vtree()` will not overwrite labels if
they already exist.

For example, to mimick the `sameline` option in vtree, you can first
generate the standard automatic labels and then replace the newline
character with a space:


```{r eval=FALSE}
## vtree::vtree(FakeData, "Severity Sex", sameline = TRUE)
vtree(FakeData, Severity, Sex) |>
  add_labels() |>
  mutate(label = gsub("\n", " ", label)) |>
  plot(lwidth = .9)
```

Arguably, that is way more code than just adding `sameline = TRUE` to the
vtree() call, but it is also way more flexible.

**`labelnode`**. This is used in vtree to replace a variable level with a
custom label. In vtree2, you should simply replace the variable levels *in
the data* before constructing the vtree.

```{r eval=FALSE}
## vtree::vtree(FakeData,"Group Sex",horiz=FALSE,
##        labelnode=list(Sex=c(Male="M",Female="F")))
FakeData |>
  mutate(Sex = dplyr::recode(Sex, M = "Male", F = "Female")) |>
  vtree(Group, Sex) |>
  plot(dir = "tb")
```

#### Targetting nodes

Prune / keep and labelling operation might want to target certain nodes by
their path. This is achieved in vtree using `tlabelnode`. In vtree2, each
node has an `ID`, a character representation of a path, and a `path` column
with the full path specified as list. You can use these to target nodes for
pruning, keeping, labelling or coloring. More: you can use your prunning or
keeping operation not to actually remove the nodes, but to just select them
for further operations. Thus, you can find a node ID, use prune to find all
nodes that *would* be removed, and then color them in red.


```{r eval=FALSE}
data(titanicNA)
vt <- vtree(titanicNA, Class, Sex, Survived)
vt |> keep(ID == "Class:1st/Sex:NA/Survived:Yes" |
           ID == "Class:2nd/Sex:NA/Survived:Yes") |>
  plot()

vt |> add_labels() |>
  add_palette() |>
  mutate(label = ifelse(ID == "Class:1st", "First class", label)) |>
  mutate(fill = ifelse(ID == "Class:1st", "red", fill)) |>
  plot()

vt |> prune(ID == "Class:2nd/Sex:NA", mark_only=TRUE) |>
  add_palette() |>
  mutate(fill = ifelse(mark != "keep", "red", fill)) |>
  plot()
```





#### Plotting

The **`horiz=TRUE`** option in vtree directs whether the tree is plotted
horizontally or vertically. In vtree2, you can specify the direction of the
plot with the `dir` parameter in `plot()`. The default is "lr" (left to
right), but you can also use "tb", "bt", or "rl".


**Changing variable labels**. In vtree, you can specify variable labels
with the `labelvar` parameter. In vtree2 you can use the column labels you
wish to see in the original data. If you have weird characters, use the
backticks.



```{r eval=FALSE}
## vtree::vtree(FakeData, "Severity Sex",
##       horiz = FALSE,
##       labelvar=c(Severity="Initial severity"))
FakeData |>
  rename(`Initial\nseverity` = Severity) |>
  vtree(`Initial\nseverity`, Sex) |>
  plot(dir = "tb")
```


#### Missing functionality

The following are not yet implemented in vtree2:

 * font styling
 * the various variable processing options, like turning a numeric variable
   into a factor with a specified number of levels

### Design principles for vtree2

 * separate frequency calculations, summary calculations, node selection and plotting
 * use graphical objects for plotting which are flexible
 * use tidyverse principles for data manipulation
 * use tidyverse syntax for specifying variables
 * vtree object inherits from tidygraph's tbl_graph, so it can be used with
   a wide variety of plotting tools, including ggraph, ggplot2, and plotly

### Practical differences


 * In vtree2, you first prepare the data with `vtree()` and then plot it with `plot()`. 
   This means one more function call, but a much more flexible plotting system.
 * Given that a vtree object is a tbl_graph, you are not limited to the
   built-in plotting functions. You can use ggraph, ggplot2, or plotly to
   create your own plots.
 * There are some automatic summary and plotting functions, but you can
   also have a very tight control about the labels by modifying the
   `label` column in the vtree object.
 * The summary_vt function can be used to calculate summary statistics for
   each node based on any additional data. It is not limited to using the
   data from which the vtree was constructed, and you can use the exposed
   statistics to shape any label you would like.
 * Same goes for colors: instead of using automatic color assignments that
   mimick these of the original vtree, you can use any color scheme you
   like by assigning values to the `color` column in the vtree object.
 * The downside of this setup is that the original vtree allows you to
   generate a beautiful plot with a single terse function call, whereas
   vtree2 is much more explicit and requires several steps. The upside is
   that you have more flexibility and more control over the final plot.
 * Using ggplot2 unfortunately means that you might need to adapt the node
   sizes and label font sizes manually. ggplot2 is not aware of the size of
   the text and plot when constructed.
 * Similarly, the geometry of the plot is chosen by the original vtree
   plotting function, whereas in vtree2 the geometry of the plot is chosen
   by the user, and plotting function tries to fit the graph inside it,
   possibly requiring the user to adapt the font and label sizes or the
   geometry.
 * The mini-language implemented in vtree is now replaced by explicit
   operations: finding nodes, labelling, selecting colors for plotting etc.

## AI disclosure

I have not used LLMs directly in writing the code or documentation.
However, I did use an LLM based chat to discuss some of the design issues
(e.g. to compare existing packages for graph representation). Also, for
code review / bug finding.

## TODO

 * ~~plotting function for patterns~~
 * make sure that node_name is used for displaying variable names
 * how is vtree actually handling the palettes?
 * ~~add the keep_children parameter to the keep() function~~, and also
   keep_na. - this is harder; because not all the NA nodes are kept, but
   only these which are on the same level as a node that is kept.
   Generally make keep behave the same as in vtree.
 * ~~legends~~
 * should the vtree() function take cases or samples?
 * maybe gridtext for the labels so we can use some basic formatting
 * faster processing of cases into trees
 * questions to NB:
   - what is the difference between seq and pattern? I don't see any at
     least on simple plots (apart from the arrows)
   - how are the NA nodes kept in pruning *exactly*?

## BUGS

 * prune(!Severity %in% c("Mild", "Moderate"), follow_only = TRUE) does not work
 * na.rm handling in prune() is not consistent, at least not with vtree's behavior
 * layout="proportional" stopped working
