---
title: "Vtree2 User Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vtree2 User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup,echo=FALSE,warning=FALSE,message=FALSE}
library(vtree2)
library(cowplot)
library(tibble)
library(dplyr)
```



# Quick start

```{r quickstart,eval=FALSE}
library(vtree2)

# example data set - same as Titanic, but with NAs
data(titanicNA)

# variables to show: Class and Survived
vt <- vtree(titanicNA, Class, Survived)

plot(vt, legend = TRUE)
plot(vt, layout = "proportional")
```

```{r quickstart_real,fig.width=10,fig.height=5,echo=FALSE,warning=FALSE,message=FALSE}
library(vtree2)

# example data set - same as Titanic, but with NAs
data(titanicNA)

# variables to show: Class and Survived
vt <- vtree(titanicNA, Class, Survived)

p1 <- plot(vt, legend = TRUE)
p2 <- plot(vt, layout = "proportional")
plot_grid(p1, p2)
```

For more information what the vtrees are and how to use them, see the [What
are Vtrees?](what_are_vtrees.html) vignette. Read on for the full user
manual.

# Vtree workflow

In `vtree2`, the workflow is split into several steps:

 * Prepare the data (outside of `vtree2`)
 * Build the vtree object with `vtree()` or `vtree_from_freqtable()`
 * (Optional) Prune, retain or select nodes with `prune()`, `retain()` and
   friends.
 * (Optional) Create summaries with `summarize_by_node()` and `fmt_label()`.
 * (Optional) Add aliases for variable names and values with `add_aliases()`.
 * (Optional) Add or modify labels with `add_labels()` and `mutate()`.
 * (Optional) Add or modify colors with `add_palette()` and `mutate()`.
 * Plot the vtree with `plot()`.

Here is an example demonstrating all these steps based on the `titanicNA`
data set. This data set is the same as the Titanic data set, but with some
missing values randomly added for demonstration purposes.


```{r workflow_example1,fig.width=10,fig.height=5}
data(titanicNA)

vtree(titanicNA, Class, Sex, Survived) |>
  # only retain the third class passengers
  retain(path == "Class:3rd") |>
  # change how variables are displayed
  add_aliases(val_alias = list(Class = c("1st" = "First",
                                    "2nd" = "Second",
                                    "3rd" = "Third"))) |>
  # add default labels, built from aliases
  add_labels() |>
  # change the labels for the missing values
  mutate(label = gsub("NA", "Missing", label)) |>
  # add colors
  add_palette(palettes = c("Greys", "Blues", "Purples"),
              na = "grey90") |>
  # change the color for Females who survived
  # mark() adds a column `mark` to the vtree object, which is TRUE for
  # nodes that match the given condition
  mark(path == "Class:3rd/Sex:Female/Survived:Yes") |>
  mutate(fill = ifelse(mark, "red", fill)) |>
  mutate(color = ifelse(mark, "white", color)) |>
  mutate(label = ifelse(mark, sprintf("**%s**", label), label)) |>
  # plotting with legend and custom margins
  plot(legend = TRUE,
       richtext = TRUE,
       margins = c(0.05, 0.05, 0.25, 0.05))
```



# Building vtree objects

## Case data

Vtrees are primarly built from *case data*: matrices or data frames in
which each row corresponds to a single sample. Each column is a data
variable and can be selected to be used and displayed on a vtree. For
example, consider the built-in data set `titanicNA`. This is the same data
set as `Titanic` from the `datasets` package, except that it is a cases
data frame and some of the values were replaced by NAs to simulate missing
data:


```{r vtree1}
data(titanicNA)
head(titanicNA)
```

As you can see, we have four variables: `Class`, `Sex`, `Age` and
`Survived`. Each of the `r nrow(titanicNA)` rows corresponds to a single
passenger.

The `vtree()` function then collects the levels of each selected variable
and counts the number of samples for each combination of levels:


```{r vtree2}
vt <- vtree(titanicNA, Class, Survived)
vt
```

Note that you select the variables that you want to consider in your vtree
- in the above example, we do not care about Sex or Age. Also note that the
variables follow the tidyverse logic - they are *data vars*, and do not
need to be quoted. If you need to use a variable name that is stored in
another variable, or if the variable names are not valid R identifiers, you
can use `tidyselect::all_of()` to select the variables.


```{r vtree3}
vars <- c("Class", "Survived")
vt <- vtree(titanicNA, all_of(vars))
```

That is one way to construct vtrees. The other is to use frequency tables.

## Frequency data

Often the data is already summarized in a frequency table, where each row
corresponds to a combination of levels of the selected variables, and there
is a column with the number of samples in which this combination occurs.

There are two functions to construct vtrees from frequency tables:
`vtree_from_freqtable()` and `cases_from_freqtable()`. The first one
constructs the vtree directly from the frequency table, while the second
one first expands the frequency table into a cases data frame which you can
use to construct the vtree with `vtree()`.

The original Titanic data set is a cross-tabulation of the four variables.
When converted to a data frame, it becomes a frequency table, with the
column `Freq` containing the number of passengers for each combination of
levels of the four variables:


```{r vtree4}
head(as.data.frame(Titanic))
cases <- cases_from_freqtable(Titanic)
vt <- vtree(cases, Class, Survived)

# or do it directly:
vt <- vtree_from_freqtable(Titanic, Class, Survived)
```

## Data preparation

The vtree object constructors (`vtree()` and `vtree_from_freqtable()`) can
only use factors and character values. They flatly refuse to process
numeric or logical columns. This is by design: numeric and logical values
are not categorical, and vtrees are meant to visualize categorical data.

Also, converting numerical or logical values to factors or character
vectors is easy. It is best done with regular R code.


## Valid percentages

One of the important concepts to know about are the *valid percentages*. In
short, a valid percentage is the number of samples with a given value of a
variable divided by the total number of samples *for which this variable is
not missing*.

If you have 50 males, 50 females and 100 samples for which
the sex is unknown, then the number of males divided by the total number of
samples may be misleading (25%); valid percentages are calculated only from
the 100 samples for which the sex is known and the proportion of males and
females is 50% each.

The `vtree()` and `vtree_from_freqtable()` functions by default calculate
valid percentages; if you want to calculate the absoluet frequencies, use
the `.vp=FALSE`argument to these functions.

Once the vtree has been constructed, this cannot change.
Any operations downstream (like selecting and pruning nodes and modifying
their labels) will not change the calculated percentages.

## Vtree objects

You can inspect the vtree object with a number of methods.

```{r vtree5}
# names of the column variables associated with the nodes
names(vt)

# available levels of the different column variables:
levels(vt)

# summary frequency data for each node variable
summary(vt)

# check whether the vtree is calculated using valid
# percentages or absolute frequencies
is_vp(vt)
```

The vtree objects contain a data frame of nodes with a number of associated
columns, which include:

 * `node_key` - a unique identifier for each node
 * `node_id` - a unique integer identifier for each node
 * `node_col` - the name of the column variable associated with the node
 * `node_val` - the value of the variable associated with the node
 * `freq` - the percentage of samples within the node that have the given
   `node_val`
 * `n` - the number of samples in the current node
 * `missing` - the number of missing samples in the current node
 * `tot_n` - the number of samples in the parent node
 * `denom` - the number of samples within the node that have a non-missing value for the
   variable associated with the node. This is equal to `tot_n` only if the tree
   was not constructed with valid percentages. Otherwise, it is equal to
   the number of samples in the parent node that have a non-missing value
   for variable `node_col`.

Some other columns may be added later with `mutate()` or with the various `add_*()` functions, for example:

 * `label` - the label to be displayed on the node
 * `fill` - the fill color of the node
 * `color` - the text color of the label
 * `x`, `y`, `width`, `height` - layout information for the node

You can inspect the columns of the vtree object with `nodecols()`, which is
just a wrapper around `colnames(as_tibble(vt))`:


```{r vtree6}
nodecols(vt)
```

You can convert the vtree object to a data frame with `as_tibble`:

```{r vtree7}
as_tibble(vt)
```

## Vtree objects as graphs

The `vtree` class is a wrapper around the `tbl_graph` class from package
`tidygraph`. It is a directed graph, where each node corresponds to a
combination of levels of the selected variables. The `tbl_graph` class
itself inherits from `igraph`, which means that you can use all of the
`igraph` functions on `vtree` objects:


```{r vtree8}
igraph::degree(vt)
```

Since the default plotting function for `igraph` objects is replaced by
`vtree2`'s own `plot.vtree()`, if you want to use `igraph`'s plotting
function, you need to convert the vtree object to a `tbl_graph` object first:


```{r vtree9,fig.width=4,fig.height=4}
vt_tbl <- as_tbl_graph(vt)
plot(vt_tbl)
```

## Tidy selection of variables

The selection of columns in `vtree()` and related functions follows the
`tidyselect` rules. You don't quote the data var names, you can use `:` to
select a range or `-` to exclude a variable. For programmatic selection,
use a character vector and `tidyselect::all_of()` to select the variables.
The following are equivalent:


```{r eval=FALSE}
vtree(titanicNA, Class, Sex, Age)
vtree(titanicNA, -Survived)
vtree(titanicNA, Class:Age)
vtree(titanicNA, all_of(c("Class", "Sex", "Age")))
```




# Pruning, retain and selecting

## Pruning and keeping

Quite often, we don't want to show the whole tree, but only selected nodes.
For this, you can use the functions `prune()` and `retain()`. Each has a
few pecularities and defaults.

**Pruning** a node means: remove the node from the tree, along all the
nodes that follow it. However, with the parameter `follow_only=TRUE`, all
the following nodes will be pruned, but the given node will be kept.

**Retaining** means that you indicate which nodes you would like to keep,
and prune all other nodes.

Both pruning and retaining take a logical expression as an argument. The
expression is evaluated in a particular context: you can use both all the
node data frame columns (i.e., data vars such as `freq`, `n` or `path`), as
well as the node variables (like `Class` or `Survived`). All this works
naturally: you can specify that you want only to see the node where 
`Class == "1st"` or all nodes where `freq > 0.5`.

```{r nodes1,fig.width=9,fig.height=9}
# all variables except for Age
vt <- vtree_from_freqtable(Titanic, -Age)
p1 <- plot(vt)
# show only 1st class passengers
p2 <- vt |>
  retain(Class == "1st") |>
  plot(legend = FALSE, lwidth =.8)
# remove nodes with low frequency or number
p3 <- vt |>
  prune(freq < .2 | n < 20) |>
  plot(legend = FALSE, lwidth =.8)
# show only combinations of factors where more than a quarter of passengers
# survived
p4 <- vt |>
  retain(Survived == "Yes" & freq > .25) |>
  plot(legend = FALSE, lwidth =.8)
plot_grid(p1, p2, p3, p4, nrow = 2,
          labels=c("Original", "Retained Class:1st",
         "Pruned freq < .2 | n < 20",
         "Retained freq > .25 & Survived:Yes"),
         label_size = 10)
```


**Valid percentages and NA nodes.** There is one pecularity about how the
NA nodes are shown. If you try to prune a node corresponding to a missing
value, it will still be shown - if the tree was constructed with the
so-called "valid percentages" (`.vp=TRUE`, default). VP means that the
denominator used to calculate frequencies on the tree was the number of
*valid* samples, i.e. excluding the missing values. Therefore, without the
NA node shown for a VP-based tree, you lack information necessary to
understand the percentages shown.

```{r nodes2,fig.width=9,fig.height=6}
# all variables except for Age
vt <- vtree(titanicNA, -Age)
p1 <- vt |>
  # this should prune the NA nodes, but doesn't
  retain(Class %in% c("1st", "2nd")) |>
  prune(is.na(Sex)) |>
  plot(legend = FALSE, lwidth=.8)
vt_novp <- vtree(titanicNA, -Age, .vp = FALSE)
p2 <- vt_novp |>
  # this one prunes the NA nodes, b/c the tree 
  # is created with .vp = FALSE
  retain(Class %in% c("1st", "2nd")) |>
  prune(is.na(Sex)) |>
  plot(legend = FALSE, lwidth=.8)
plot_grid(p1, p2, nrow = 1,
          labels = c("VP",
                     "No VP"))
```

**Targetting nodes.** Quite often, you want to pick one particular node and
either prune it or select only that node for plotting.
The `path` column contains a human-readable specification of the node which
makes picking a certain node easy. For example, if you want to keep only
the node corresponding to males from 3rd class (and the following nodes),
you can ask for nodes where path is equal to `"Class:3rd/Sex:Male"`:

```{r nodes3,fig.width=8,fig.height=4}
vt <- vtree_from_freqtable(Titanic)
p1 <- plot(vt, legend = FALSE, dir="tb", show_root = FALSE)
p2 <- vt |>
  retain(path == "Class:3rd/Sex:Male") |>
  plot(legend = TRUE, dir="tb", show_root = FALSE)
plot_grid(p1, p2, nrow = 1, labels = c("Original tree", "Retained node"))
```

Note that the legend shown on the left in the above plot still contains all
the other variable levels, even if you can't see them on the tree. This is
on purpose: the stat summaries in the legend correspond to the whole data
set on which the tree was built, and not only the portion that is shown in
the tree.

## Finding and marking

Sometimes rather than change the visibility of selected nodes by pruning or
retaining, you might want to change their color or label instead. For this,
you can use `find_nodes()` to get a logical vector indicating which nodes match a
certain condition or `mark()` to produce a vtree object with a new column,
`mark`, holding that logical vector.


```{r fig.width=4,fig.height=4}
vt <- vtree(titanicNA, Class, Survived)

# which nodes have a low frequency?
mask <- find_nodes(vt, freq < 0.2)
as_tibble(vt) |> filter(mask)

# mark nodes with low frequency with pink
# all other nodes will be white
vt |> mark(freq < .2) |>
  mutate(fill = ifelse(mark, "pink", "white")) |>
  plot()
```






# Adding column and value aliases

If you prefer to see a different name for the variable or its values on the
plot, there are basically two approaches.

First, you can just recode or rename the variables in the data from which
you build the vtree. This has an advantage: you ensure that the labels
you see on the plot are the same as in the original data. For example, you
can do the following:


```{r aliases1,fig.width=4,fig.height=4}
cases <- as_tibble(Titanic) |>
  mutate(Class = recode(Class, "1st" = "First", "2nd" = "Second",
                        "3rd" = "Third")) |>
  cases_from_freqtable(.freq_col = "n")

vtree(cases, Class, Survived) |> plot()
```

However, some times what you want to see on the plot is not convenient to
work with in the data, for example if you want to show a longer description
of the possible values, add Unicode characters or markdown / HTML
formatting for rich text labels. In this case, you can specify an alias for
either using the `col_alias` and `val_alias` arguments of `add_aliases()`.
Not all variables or all levels (values) need to be specified, only those
you want to change:

```{r aliases2,fig.width=8,fig.height=4}
col_alias <- list(Class = "Passenger\nClass",
                  Sex = "Gender",
                  Age = "Age\nGroup",
                  Survived = "Survival\nStatus")

val_alias <- list(NAs = "Missing",
                  Class = c("1st" = "First",
                             "2nd" = "Second",
                             "3rd" = "Third"))

vtree(titanicNA, Class, Survived) |>
  add_aliases(col_alias = col_alias,
             val_alias = val_alias) |>
  plot(legend = TRUE, dir="tb", show_root = FALSE,
       margins = c(0.05, 0.05, 0.05, 0.10))
```

The special `NAs` name is used for all missing values, regardless of the variable.

Unnecessary gory details:

 * `add_aliases()` and `add_labels()` always ensure that columns
   `col_alias` and `val_alias` are present in the vtree object.
 * If no custom format is specified, these two columns will be identical to
   the `node_col` and `node_val` columns, respectively.
 * In addition, the alias information is stored in the `alias`
   attribute of the vtree object, which is subsequently used
   by `plot()` to create the legend.

# Adding and modifying labels

Labels to be plotted are taken from the `label` column of the vtree object.
If this column is missing, `plot()` calls `add_labels()` to add default
labels. For custom labels there are three routes:

 * directly add a label column to the vtree object with `mutate()`
 * use `add_labels()` with a custom format
 * first use `add_labels()`, then modify the label column with `mutate()`

## Directly creating a label column with `mutate()`

Since the `mutate()` function is called in the context of the vtree node
data frame, you can use any of the columns of the vtree object directly.
For example, you can show the `node_key` column, a unique identifier for
each node.

```{r labels1,fig.width=8,fig.height=4}
vtree(titanicNA, Class, Survived) |>
  mutate(label = paste0("node_key:\n", node_key)) |>
  plot(legend = FALSE)
```

### Using `add_labels()` to add default or custom labels

The `add_labels()` function can produce two types of default labels:
"simple" and "long", chosen through the `template` argument:

```{r labels2,fig.width=8,fig.height=4}
vt <- vtree(titanicNA, Class)
p1 <- add_labels(vt) |> plot() # template = "simple"
p2 <- add_labels(vt, template = "long") |> plot()
plot_grid(p1, p2, nrow = 1)
```

These labels are derived directly from columns of the vtree object:
`col_alias`, `val_alias`, `freq` and `n`. If no aliases are set, then
`col_alias` is the same as `node_col`, and `val_alias` is the same as
`node_val`.

It is quite normal to call `add_labels()` multiple times, for example to
conditionally add labels. The following makes simple default labels and
adds a more informative label only to the leafs on the tree:

```{r labels2b,fig.width=4,fig.height=7}
vt <- vtree(titanicNA, Class, Survived)
mask <- pull(vt, leaf) # leaf is a logical vector
                       # indicating whether a node is a leaf node
vt |>
  add_labels(template = "sameline") |>
  add_labels(mask = mask, template = "long") |>
  plot()
```

## Using custom formatting

By default, `add_labels()` produces simple node labels containing the
associated variable value, number of cases and percentage within the
parent node. This can be customized by one of the following:

 * choose a different `template` parameter: `short` (default),
 `sameline` (same as short, but on one line) or `long` (with variable
 names). The templates all reasonably handle NA nodes and root node.
 * use a [glue::glue()] syntax for the parameters `fmt`, `fmt_na` and
 `fmt_root`, where variable names are put in curly
 braces. The variable names are the same as column names of the node
 data frame of the vtree object, plus `pct` and `f` (see below). The
 three parameters will be used to generate regular, NA-nodes or root
 node labels, respectively.
 * use an arbitrary R expression (parameter `expr`) which is evaluated with
 [rlang::eval_tidy()] in the context of the nodes data frame of the
 vtree object.

Both the glue format syntax and the arbitrary expression syntax can use
any column name which is already present in the nodes data frame,
including:

 * `freq`, the frequency for a node
 * `n`, number of samples of a node
 * `col_alias`, the alias for the column/variable associated with a node
   (default same as node_col, but can be modified
   by providing a `var_alias` column in the vtree)
 * `val_alias`, the alias for the value of the variable associated with a node
   (default same as node_val, but can be modified 
   by providing a `val_alias` column in the vtree)
 * `node_col`, name of the variable associated with a node
 * `node_val`, value of the variable associated with a node
 * plus whatever new columns you have added to the vtree with mutate().

You can list the columns in the node data frame of a vtree object with
`nodecols(vt)` (or simply `colnames(as_tibble(vt))`).

In addition, `add_labels()` provides two additional, preformatted
values:

 * `pct`, percentage rounded to the specified number of digits (the
 `digits` parameter)
 * `f`, equal to pct / 100 (so if the percentage is rounded with 0
 digits after decimal point, `f` will have two digits after decimal
 point).

Here is a formatting example with the `fmt_*` parameters.

```{r labels3,fig.width=4,fig.height=4}
vt <- vtree(titanicNA, Class, Sex) |>
  add_labels(fmt = "{node_col}: {node_val}\nfreq={f} ({pct}%)\nN={n}",
             fmt_na = "{node_col}: Missing\nN={n}",
             fmt_root = "All data:\nN = {n}")
plot(vt, legend = FALSE)
```

## Using the `expr` parameter

If you need even finer control, you can use the `expr` parameter which
takes any R expression and evaluates it with `[rlang::eval_tidy()]` in the
context of the nodes data frame of the vtree object (plus the variables
`pct` and `f`). 


```{r labels_expr,fig.width=5,fig.height=6}
vt <- vtree(titanicNA, Class, Survived) |>
  add_labels() |> # add default labels
  add_labels(expr =
             ifelse(node_col != "Survived",
               label,
               paste0(
                 ifelse(freq > .5, "Many ", "Few "),
                 ifelse(node_val == "Yes", "survived", "perished"),
                 "\n",
                 label
                 )
               ))
             
             
plot(vt, legend = FALSE)
```


## Using a mask

You can specify a mask with `add_labels()` - a logical vector with the same
length as the number of nodes - to indicate which nodes should have their
labels modified. For example, you may want a different information for the
leaf nodes as for other nodes:

```{r labels_mask,fig.width=5,fig.height=4}
vt <- vtree_from_freqtable(Titanic, Class, Sex, Survived) |>
  retain(path == "Class:1st") # only keep the first class passengers
mask <- find_nodes(vt, leaf) # leaf is a logical vector
add_labels(vt, mask = mask, template = "long") |>
  add_labels(mask = !mask, fmt = "{node_val}") |>
  plot(legend = FALSE, dir="tb", show_root = FALSE,
       lwidth = 0.7, lheight = 0.5)
```

## Parameter precedence with `add_labels()`

If `expr` is not NULL, it will be used for all labels chosen by the
mask.

If `fmt` is NULL, the selected template will be used.
If `fmt_na` is NULL and `fmt` is not NULL, then `fmt_na` will be `fmt`,
otherwise the selected template will be used. Same for `fmt_root`: first
`fmt`, if defined, otherwise the template.

`fmt_na` is used for NA values only if `is_vp(vtree)` is `TRUE`; this is
because for a vp tree the NA value percentages are meaningless.


# Creating summaries

One of the more useful aspects of using vtrees is that you can partition
the data using a set of variables and then show summary information *for
another variable*. For example, consider the ToothGrowth data set,
containing two categorical variables – dose and type of supplement – as
well as numeric result. We can use the categorical variables for data
splits and ask what the average tooth growth is at the different nodes.

```{r summaries1,fig.wilth=8,fig.height=5}
# make sure that dose is a factor
tg <- ToothGrowth |>
  mutate(dose = factor(paste("Dose", dose)))

vt <- vtree(tg, supp, dose)

library(glue)
sm <- summarize_by_node(tg, vt, len) |>
  fmt_label()

vt |>
  add_aliases(col_alias = c(supp = "Supplement",
                            dose = "Dose (mg/day)")) |>
  add_labels(template = "sameline") |>
  add_labels(fmt = "{label}\n{sm}") |>
  plot(lwidth=.75)
```

The above example demonstrates the main workflow of generating summaries:

 * you need a cases data frame to generate summaries
 * generate summaries in a separate step
 * use `summarize_by_node()` to compute, `fmt_label()` to format

It is also possible to include graphical summaries; see the section "Inset
plots" for more details.

## Summaries with `summarize_by_node()`

Each row of the data frame returned by `summarize_by_node` corresponds to
one row of the nodes data frame in the vtree, hence the name
of the function.

The columns in the data frame returned by `summarize_by_node` depend on
whether the data is categorical or numeric. For numeric data, as above, the
columns include mean, sd, n, quantiles 1, 3, and median, iqr, as well as
the number of valid and missing samples. 

```{r summaries2}
# numerical data
sm <- summarize_by_node(tg, vt, len)
head(sm)
```

For categorical data, the function returns a data frame including total
number of observations, number of valid, missing and unique observations, a
list column containing the counts for each level of the variable chosen, as
well `denom` (denominator) and the calculated frequencies of individual
levels.

Here we return to the Titanic example and ask, what was the fraction of
women who survived or not for the different Classes. We use the `titanicNA` data
set, which contains some randomly introduced missing data.

```{r summaries3}
# categorical data
vt <- vtree(titanicNA, Class, Survived)

sm <- summarize_by_node(titanicNA, vt, Sex)
head(sm)
```

As when calculating the vtree, for calculating a frequency of a level for a
variable in the above summary (e.g. the frequency of females among 1st
Class passengers), we can choose whether we use valid percentages or not.
That is, we have to decide whether we calculate the fraction as the
fraction of *valid* samples, or the fraction of *total* samples at this
node. In the above, the total fraction is $0.458 = 120/262$ - we see that
by looking at the `denom` (denominator) column of the table above.

By default, `summarize_by_node()` uses the same procedure as was used for
the vtree provided as parameter. However, you may choose to control this
behavior with the `vp` parameter:


```{r summaries4}
# categorical data
vt <- vtree(titanicNA, Class, Survived)

sm <- summarize_by_node(titanicNA, vt, Sex, vp=FALSE)
head(sm)
```

The denominator above is equal to column `n`, total number of cases per
node.

## Formatting statistics with `fmt_label()`

The function takes the output from `summarize_by_node()` and converts it to
a character vector of labels. By default, the function looks into the
`type` column of the data frame to determine whether the summarized
variable is categorical or numeric, and serves either a categorical or
numerical summary of the data.

The output can be customized either with glue format strings (argument
`fmt`) or R expressions (argument `expr`). The following are equivalent:


```{r}
vt <- vtree(titanicNA, Class, Survived)

sm <- summarize_by_node(titanicNA, vt, Sex, vp=FALSE)

txt1 <- fmt_label(sm, fmt="{missing} {n} {Male_freq}", digits=2)
txt2 <- fmt_label(sm, expr=glue("{missing} {n} {round(Male_freq, digits=2)}"))

all(txt1 == txt2)
```

There are two differences between `fmt` and `expr`:

 * `fmt` uses rounded versions of numeric columns. Each numeric column is
   rounded with the specified number of digits (argument `digits`).
 * `expr` allows any kind of R expression, including conditional
   expressions with `ifelse`, `if_else` and `case_when`, and string
   formatting with `sprintf`, `glue` etc.

## Custom statistics with `vtree_apply()`

Sometimes you need to calculate some per-node statistics which goes beyond
simple descriptive statistics.

`vtree_apply()` goes through nodes in the vtree object, and for each
object, finds the matching rows in the `cases` data frame. Then it calls
the user-provided function and calls it with some arguments.

The arguments are also configurable. By default, the function is called
with the subset of the original `cases` data frame which corresponds to the
given node. For example, we can calculate how many samples there are for a
node:

```{r summaries5}
cases <- cases_from_freqtable(Titanic)
vt <- vtree(cases, Class, Survived)

vtree_apply(cases, vt, nrow) |> unlist()

# same as:
vt |> pull(n)
```

However, you can configure that also to provide e.g. the logical selection
vector. Say, we want, for each node, determine whether the male:female
ratio in that node is significantly different from the rest of the data.


```{r vtreapply1}

test_func <- function(sel) {
  # root node includes all samples
  if(sum(sel) == length(sel)) {
    return(NA)
  }

  chisq.test(table(cases$Sex, sel))$p.value
}

pvals <- vtree_apply(cases, vt, test_func, .args="sel") |> unlist()
pvals_l <- format.pval(pvals, digits=2)

# color the nodes by p-value
logp <- floor(-log10(pvals))
pal <- colorRampPalette(c("white", "red"))(101)

# use simpler summaries 
sm <- summarize_by_node(cases, vt, Sex) |>
  fmt_label(fmt = "Females: {Female_freq * 100}", digits=2)

vt |> add_labels(suffix = sm) |>
  add_labels(fmt = "{label}\n{pvals_l}",
             fmt_root = "{label}") |>
  mutate(fill = pal[logp + 1]) |>
  plot()
```




# Adding and modifying colors

There are basically three ways of modifying colors shown on the plot and
ont the legend:

 * use `add_palette()` with `palette` argument to assign a palette to each
   variable. You can use the `RColorBrewer` palettes for that. This will
   change the colors both on the plot and on the legend.
 * use `add_palette()` with `var_levels` and `var_colors` arguments to
   assign colors directly to each variable and variable level. This will
   change the colors both on the plot and on the legend.
 * change the columns `fill` and `color` in the nodes data frame directly
   with `mutate()` (possibly after generating some defaults with
   `add_palette()`). This will change the colors only on the plot, but not
   on the legend.

On the plot, the `fill` and `color` columns determine how each node is
colored; the palettes stored in the `palette` attribute of the vtree object
determine how the legend is colored.

If the `fill` column is missing from the vtree object, `plot()` will call
`add_palette()` to assign fill colors automatically. If `color` is missing, but
`fill` is present, then white or black will be chosen automatically
depending on the contrast with the fill color.

## Modifying colors with `mutate()`

It is not necessary to call `add_palette()` if you just want to change the
color of the text and fill on the nodes, without changing the legend:

```{r colors1,fig.width=8,fig.height=4}
pal <- colorRampPalette(c("white", "steelblue"))(101)

p1 <- vt |>
  mutate(fill = pal[round(freq * 100) + 1]) |>
  plot(lwidth=.6)

p2 <- vt |>
  mutate(abs_freq = n / max(n)) |>
  mutate(fill = pal[round(abs_freq * 100) + 1]) |>
 plot(legend = TRUE, lwidth=.6)

plot_grid(p1, p2)
```

Colors can be also assigned automatically based on names of RColorBrewer
palettes. There is a default order of palettes for the variables, but you
can override it with the `palettes` argument of `plot()` or
`add_palette()`.

## Adding palettes with `add_palette()`

The `add_palette()` function not only sets the `fill` and `color` columns, but 
also stores the palette in the attribute `palette` of the vtree object.
This palette in turn is used to create color keys on the legend. If that
attribute is missing, the legend will be black and white.


# Adding layout with `add_layout()`

## Basic layout configuration with `plot()`

Layouts are added with the `add_layout()` function, which you can call
directly on the vtree object, resulting in a new vtree object with
additional columns for the layout. However, if a vtree without a layout is
passed to `plot()`, then `plot()` will call `add_layout()` automatically.

Currently, there are four layouts implemented: "regular", "flushed_left",
"flushed_right" and "proportional". The default layout is "regular", which
simply shows the tree structure with all nodes having the same sizes. The
"flushed_*" layouts are similar, but the nodes are always flushed to one side
of the plot.

The proportional layout shows the nodes with sizes proportional to the
number of observations in that node.


```{r layouts2,fig.width=8,fig.height=8}
# the default column name for .freq_col is "Freq", same as in Titanic,
# so no need to specify it here
vt <- vtree_from_freqtable(Titanic, Class, Sex, Survived)
p1 <- plot(vt) # layout regular
p2 <- plot(vt, layout = "proportional")
p3 <- plot(vt, layout = "flushed_left", dir="tb")
p4 <- plot(vt, layout = "flushed_right", dir="tb")
plot_grid(p1, p2, p3, p4, nrow = 2)
```

With the `layout_func` argument, you can specify a custom function to
calculate the layout yourself. For example, this function might call
`add_layout()` first and then adjust the calculated positions of the nodes
in some way. See the documentation for `add_layout()` for details.

Layouts will be automatically created by `plot()` if they are missing. The
parameters passed to the plot give some basic control over the layout:
`dir` determines the direction of the tree, `show_root` determines whether the
root node is shown, and `lwidth`/`lheight` determine the width and height
of the nodes relative to the available space.


```{r layouts1,fig.width=10,fig.height=5}
vt <- vtree_from_freqtable(Titanic, Class, Survived)
p1 <- plot(vt)
p2 <- plot(vt, dir = "tb", show_root = FALSE, lwidth = 0.8, lheight = 0.3)
plot_grid(p1, p2, nrow = 1)
```

The `plot()` function simply calls `add_layout()` to add the layout to the
vtree object with the specified parameters. Alternatively, you can call the
function `add_layout()` directly on the vtree object. The resulting vtree
contains the additional columns `x`, `y`, `width` and `height` for the
nodes and `x1`, `y1`, `x2`, `y2` for the edges. You can then modify these
columns directly.

## Directly modifying the layout

In the following example we will modify a layout with pruned nodes. We will
move the nodes on the right (leafs) hand up a bit upwards.

```{r layouts3,fig.width=5,fig.height=5}
vt <- vtree_from_freqtable(Titanic, Class, Survived) |>
  prune(Class %in% c("1st", "2nd"), follow_only = TRUE) |>
  add_layout() |>
  mutate(y = ifelse(leaf, y + height/2, y))

# we need to modify the edges as well, but for that we need information
# from the nodes
nodes <- as_tibble(vt)

vt <- mutate(vt, x1 = nodes$x[from], y1 = nodes$y[from],
                 x2 = nodes$x[to] - nodes$width[to]/2, y2 = nodes$y[to],
             .edges = TRUE)

plot(vt)
```

Note how we mutate the edges with the `.edge = TRUE` argument. This is
different from `tidygraph`, where the object (tree) can either be in
activated "node" or "edge" mode. In `vtree2`, the object is always in
"node" mode (because it is more practical), but on the rare occassion that
you need to access the edges, you can.

## Fine control with `add_layout(varspace = ..., varsize = ...)`

The `add_layout()` function has additional parameters: `varspace` and
`varsize`. These can assign the amount of relative space available to the
node depending on the variable it is associated with. For example, we might
want to show a long summary which takes a lot of space for the leafs, but
only short labels for the internal nodes. You can tune the relative node
sizes with these two parameters.

```{r layouts4,fig.width=8,fig.height=2}
cases <- cases_from_freqtable(Titanic)
vt <- vtree(cases, Class, Sex, Survived) |>
  prune(Class == "Crew")

sm <- summarize_by_node(cases, vt, Age) |>
  fmt_label()

# get the logical vector indicating whether the node is a leaf
mask <- pull(vt, leaf)

vt |>
  add_labels() |>
  add_labels(mask = mask, fmt = "{label}\n{sm}") |>
  add_layout(dir="tb", lheight=.8,
             varspace=c(root=1, Class=1,Sex=1,Survived=4)) |>
  # legend = FALSE: not even column names on the margin
  plot(legend = FALSE)
```

Note that both `varspace` and `varsize` describe the node sizes along the
axis of the plot. That is, for horizontal layouts, they control the width
of the nodes; for vertical layouts, they control the height of the nodes.

# Plotting

Objects of type vtree can be directly plotted with the `plot()` method. If
the labels, colors and layouts are missing, then plot adds some automatic
defaults by running (essentially) 
`vtree |> add_labels() |> add_colors() |> add_layout()` before plotting. 
Some of the arguments for these functions
can be passed directly to the `plot()` function.

## Layouts

Parameters `layout`, `dir`, `show_root`, `lwidth`, `lheight` are passed to
`add_layout()` to control the layout of the tree. See above for details.

## Colors, fill colors and labels

## Legends with summaries

The legend shows a summary for each variable underneath or next to the
part of the tree which corresponds to this variable. Variable summaries are
calculated when vtree is constructed and stored in the `summary` attribute
of the vtree object. You can print them with `summary(vtree_object)`:

```{r plot_summaries1}
vt <- vtree_from_freqtable(Titanic, Class, Sex, Survived)
summary(vt)
```

The summaries will not change if you prune the tree, since they are tied to
the data and not the tree structure:


```{r plot_summaries2,fig.width=8,fig.height=4}
mar <- c(0.05, 0.05, 0.33, 0.05)
p1 <- plot(vt, legend = TRUE, lwidth=.8, margins = mar)
p2 <- retain(vt, path == "Class:1st") |>
  plot(legend = TRUE, lwidth=.8, margins = mar)
plot_grid(p1, p2, nrow = 1)
```


## Other `plot()` arguments

**`lwidth`, `lheight`** - label width and height relative to available space.
The layout functions calculate the available space for each node, and then
these parameters determine how much of that space is used for the label.

For the proportional layout, `lheight` is ignored, since the height of the
node is determined by the number of observations in that node.

**`margins`** - a numeric vector of length 4, specifying the top, right,
bottom and left margins as fractions of the total device space. So, for
example, for 10% margins on all sides, use `margins = c(0.1, 0.1, 0.1, 0.1)`.

**`show_root`** - if TRUE (default), show the root node (total population).

**`legend`** - if FALSE or "none", not even minimal variable legend is
shown on the plot. If "tiny" (default), the variable (column) names are
shown on the margin. If "full" or TRUE, the variable names and the
summaries are shown on the margin.

```{r fig.width=12,fig.height=4}
vt <- vtree_from_freqtable(Titanic, Class, Survived)
p1 <- plot(vt, legend = FALSE)
p2 <- plot(vt) # or legend = "tiny"
p3 <- plot(vt, legend = TRUE) # or legend = "full"
plot_grid(p1, p2, p3, nrow = 1)
```

**`dir`** direction of the tree. One of "lr" (left to right), "rl"
(right to left), "tb" (top to bottom), "bt" (bottom to top).
Default is "lr".

```{r fig.width=20,fig.height=5}
p1 <- plot(vt) # same as dir = "lr"
p2 <- plot(vt, dir = "tb")
p3 <- plot(vt, dir = "bt")
p4 <- plot(vt, dir = "rl")
plot_grid(p1, p2, p3, p4, nrow = 1)
```

**`fontsizes`** Font sizes for the various labels are automatically fit to
the available space, but sometimes you might manually adjust them. The
`fontsizes` is a named list with the following optional fields: 
 * `nodes`: node labels (default: "fixed" for regular layout,
            "adaptive" for proportional layout)
 * `var_labels`: variable labels (default: "fixed")
 * `legend_labels`: for variable levels when `legend=TRUE` (default: "fixed")

Each element can be either a number, in which case the given font size is
directly used, or either "fixed" or "adaptive". When "fixed" is used, then
optimal font size is calculated for all objects within the group, and then
the minimal font size is used for all objects. When "adaptive" is used, then each
object is fit separately, and the font sizes can be different for each object.

**`padding`**: you can try to tweak this parameter if the padding of the
label nodes or legend is not to your liking. `vtree2` attempts to
dynamically fit the text inside of the nodes, but that does not always
works. The parameter is interpreted as the desired fraction of the
width/height of the label to be used as padding, but the effective padding
depends on the actual device size and geometry.

**`lwd`** line width for use with plotting. Unfortunately, the default line
width is not relative to the device size, but is fixed to an absolute
value; therefore, for small devices the lines may appeary too thick.

## Rich text labels

It is possible to use simple HTML or markdown formatting in the labels and
then use `richtext=TRUE` in the `plot()` function to render the labels with
the formatting. The plot rendering is noticably slower, but it allows for
more flexibility in the label formatting.

```{r richtext1,fig.width=4,fig.height=4}
vt <- vtree_from_freqtable(Titanic, Class, Survived) |>
  add_labels() |>
  mark(path == "Class:1st/Survived:No") |>
  mutate(label = ifelse(mark, paste0('**', label, '**'), label)) |>
  mark(path == "Class:2nd/Survived:No") |>
  mutate(label = ifelse(mark,
                        paste0('<span style="color:red">', label, '</span>'),
                        label))

plot(vt, richtext = TRUE, legend = TRUE)
```

You can also include the rich text formatting in the aliases for the
variable names and values, which will then be used both in the labels and
in the legend.

```{r richtext2,fig.width=8,fig.height=4}
vt <- vtree_from_freqtable(Titanic, Class, Sex, Survived)

# we can use aliases through add_labels() to change the formatting of the
# variable names and values
cols <- list(Class = "**Class**", Sex = "**Sex**",
             Age = "**Age**", Survived = "**Lived**")
vals <- list(Class = c("1st" = "1<sup>st</sup>", "2nd" = "2<sup>nd</sup>",
                  "3rd" = "3<sup>rd</sup>"),
            Sex = c("Male" = "**<span style='color:blue'>Male</span>**",
                    "Female" = "**<span style='color:lightblue'>Female</span>**"))
vt <- add_aliases(vt, col_alias = cols, val_alias = vals)

plot(vt, richtext = TRUE, legend = TRUE)
```


# Patterns


# Inset plots

## Images and other graphical objects

It is possible to add any kind of "grob" - graphical object - to the nodes
description. This includes not only bitmap images, but also `ggplot2`
plots, which can be easily converted to a grob with the function
`ggplot2::ggplotGrob()`.

The important thing to remember is that the inset graphics (1) must be a
grid package graphical object ("grob") (2) you should use the
`add_graphics()` function to add it to the vtree.

In the following example, we will use three images from Wikipedia
corresponding to the three species of iris flowers from the famous Fisher
data set. We will show on the vtree how many of each of the species have
long petals, and we will illustrate the Species nodes with the
corresponding images.

```{r inset1,fig.width=7,fig.height=4}
library(dplyr)
library(purrr)

# images are included with the package
iris_imgs <- list(
versicolor="images/500px-Blue_Flag,_Ottawa.jpg",
setosa="images/500px-Irissetosa1.jpg",
virginica="images/500px-Iris_virginica_2.jpg")

get_grob <- function(img) {
  print(img)
  tf <- system.file(img, package="vtree2")
  print(tf)
  img <- jpeg::readJPEG(tf)
  # this function converts a bitmap
  # to a grid::grob object
  grid::rasterGrob(img)
}

iris_grobs <- lapply(iris_imgs, get_grob)

vt <- iris |>
  mutate(Long_Petals = as.character(Petal.Length > 4)) |>
  vtree(Species, Long_Petals) |>
  add_labels() |>
  mutate(label = ifelse(leaf, label, paste0("Iris\n", node_val))) |>
  add_palette(palettes = c("Greens", "Blues")) |>
  add_layout(dir="tb", show_root=FALSE, lwidth=.9, lheight=.8,
             varspace=c(Species=4,Long_Petals=1))

# this makes the grob list match the node list
iris_grobs <- iris_grobs[ pull(vt, "node_val") ]

vt |>
  add_graphics(iris_grobs) |>
  plot(margins=c(.05, .05, .05, .2))
```

The `add_graphics()` function has parameters which control the layout of
the node: `side` determines the side of the node where the graphics is
shown (`l` for left, `t` for top etc.), and `frac` determines the fraction
of space (vertically or horizontally) which is occupied by the graphics:


```{r inset1b,fig.width=7,fig.height=4}
vt |>
  add_layout(dir="tb", show_root=FALSE, lwidth=.9, lheight=.4,
             varspace=c(Species=2,Long_Petals=1)) |>
  add_graphics(iris_grobs, side="l", frac=.5) |>
  plot(margins=c(.05, .05, .05, .2))
```




## Inserting ggplot2 objects

Here, we choose only the "Sex" and "Survived" nodes for the construction of
the vtree. However, we make a ggplot object for each of the leaf nodes
("Survived:Yes" and "Survived:No") and insert the resulting graphics object
into the node.

For this, we create the `grob` column and convert the ggplot2 to a grob
with `ggplot2::ggplotGrob()`. The `grob` column is a list-column, so we need to
manipulate it carefully. Also, we need to set an NA value for the non-leaf
nodes.

```{r insets2,fig.width=12,fig.height=4}
# first, a tree with simplified labels on the leaf nodes
cases <- cases_from_freqtable(Titanic)
vt <- vtree(cases, Sex, Survived)

library(ggplot2)

# minimalistic ggplot showing how many passengers were in each class
# of the data frame df.
clasplot <- function(df) {
  ggplot(df, aes(x = Class, fill = Class)) +
    geom_bar() +
    scale_fill_discrete(drop=FALSE) +
    scale_x_discrete(drop=FALSE) +
    ylim(0, max(table(cases$Class))) +
    theme_minimal() +
    # remove labels, legend and axes
    theme(legend.position = "none",
          axis.title = element_blank(),
          axis.ticks = element_blank())
}

# use vtree_apply to produce plots
plots <- vtree_apply(cases, vt, FUN = clasplot)
# convert them to grid::grob
grobs <- lapply(plots, ggplot2::ggplotGrob)

# we add layout manually, so we can control it better
# this varspace arg means: reserve 4x as much space for the Survived nodes
# as for the Sex nodes
vt <- vt |> 
  add_layout(dir="tb", show_root=FALSE, lheight=.8,
             varspace=c(Sex=1,Survived=3)) |>
  mutate(shape = ifelse(leaf, "rectangle", "roundrectangle")) |>
  add_graphics(grobs, condition = leaf)

plot(vt)
```

The `condition` parameter of `add_graphics()` is evaluated in the context
of the nodes data frame, and one column in that data frame is `leaf`, which
contains a TRUE value if the node is a leaf. Therefore `condition = leaf`
means: include the grobs only for the nodes that are leafs. You can also
use the `mask` parameter for this purpose.

There is one issue with such plots: unlike `vtree2`, `ggplot2` does not fit
font sizes automatically to the available space. If your image is too
small, bad things will happen:


```{r fig.width=5,fig.height=2.5}
plot(vt)
```

## Graphical summary example

Below, we will create a graphical summary for a data set. We will use the
`ToothGrowth` data set, which contains two categorical variables
(supplement type and dosage) and the resulting length of odontoblasts in
guinea pigs. The vtree will be created using the supplement type and dosage
as variables, and we will make a graphical summary for the third variable,
length.


```{r gsum}
# first, some recoding
tg <- ToothGrowth |>
  mutate(dose = paste("Dose", dose))

vt <- vtree(tg, supp, dose) |>
  add_aliases(col_alias = c(supp = "Supplement type",
                            dose = "Dosage"))
plot(vt)
```

The graphical summary should be on leaf nodes only. It should show how the
data in that particular node compares to the remaining nodes.

For this, we need a modified `tg` data frame - one which assigns a leaf
node id to each of the samples. We can do that with the `vtree_apply`,
combined with `purrr`'s `imap_dfr` (you could also use `Reduce` and
friends). `vtree_apply` calls the provided function once per node with the
subset of the data frame that corresponds to that node; we will use it to
partition the data.

In addition, we will call vtree_apply only for the leaf nodes. That will
ensure that each sample is shown only once in the resulting data set.

```{r gsum2}
mask <- find_nodes(vt, leaf)
chunks <- vtree_apply(tg, vt, \(x) x, .mask = mask)

# the chunks correspond to the nodes
names(chunks)
as_tibble(vt)$node_key[mask] # same!

# the modified tg data frame
tg_mod <- imap_dfr(chunks, \(ch, nm) {
                     ch[["node_key"]] <- nm
                     ch })
```

Now that we have the input data frame for ggplot2, we can proceed with
constructing the plots. First draft shows how the plot should look like:

```{r gsum3,fig.width=4,fig.height=4}
library(ggplot2)
# let us select one node to mark on the plot
df <- tg_mod |>
  mutate(selected = ifelse(node_key == "node_4", "yes", "no"))

ggplot(df, aes(x = node_key, y = len, fill=selected)) +
  geom_violin() +
  geom_boxplot(width=.1, fill="white") +
  scale_fill_manual(values = c(yes="red", no="grey")) +
  theme_void() +
  theme(legend.position = "none")
```

The void theme and removing the legend make the graphic better readable
when it is squeezed into a tree node.

We just need to generate this plot for each leaf node, converting it to a
grob.


```{r gsum4}
plotfunc <- function(node, tg_mod) {
  df <- tg_mod |>
    mutate(node_key = factor(node_key, levels = names(chunks))) |>
    mutate(selected = ifelse(node_key == node, "yes", "no"))

  p <- ggplot(df, aes(x = node_key, y = len, fill=selected)) +
    geom_violin() +
    geom_boxplot(width=.1, fill="white") +
    scale_fill_manual(values = c(yes="red", no="grey")) +
    theme_void() +
    theme(legend.position = "none")
  ggplotGrob(p)
}

grobs <- map(names(chunks), plotfunc, tg_mod)
names(grobs) <- names(chunks)
```

All that remains to do is 1) assign the grobs to the vtree object, 2)
figure out how to best configure the layout.


```{r gsum5,fig.width=12,fig.height=4}
grobs <- grobs[ pull(vt, node_key) ]


vt |>
  add_palette(palettes = c("Purples", "Blues")) |>
  add_labels(fmt = "{node_val}", fmt_root = "{n}") |>
  add_layout(dir = "tb", varspace = c(root = 1, supp=1, dose=2),
             varsize = c(root = .4, supp=.4, dose=.9),
             lwidth = 1,
             lheight=.9) |>
  add_graphics(grobs) |>
  plot(legend=FALSE)
```


An alternative representation where we first split the tree by dosage and
then by supplement type allows to better compare the effects of the
supplement types:


```{r gsum6,fig.width=12,fig.height=4}
vt <- vtree(tg, dose, supp) |>
  add_aliases(col_alias = c(supp = "Supplement type",
                            dose = "Dosage"))
mask <- find_nodes(vt, leaf)
chunks <- vtree_apply(tg, vt, \(x) x, .mask = mask)

# the chunks correspond to the nodes
names(chunks)
as_tibble(vt)$node_key[mask] # same!

# the modified tg data frame
tg_mod <- imap_dfr(chunks, \(ch, nm) {
                     ch[["node_key"]] <- nm
                     ch })
grobs <- map(names(chunks), plotfunc, tg_mod)
names(grobs) <- names(chunks)
grobs <- grobs[ pull(vt, node_key) ]

vt |>
  add_palette(palettes = c("Purples", "Blues")) |>
  add_labels(fmt = "{node_val}", fmt_root = "{n}") |>
  add_layout(dir = "tb", varspace = c(root = 1, supp=2, dose=1),
             varsize = c(root = .6, supp=.9, dose=.6),
             lwidth = 1,
             lheight=.9) |>
  add_graphics(grobs) |>
  plot(legend=FALSE)
```





