---
title: "Vtree2 User Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vtree2 User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup,echo=FALSE,warning=FALSE,message=FALSE}
library(vtree2)
library(cowplot)
library(tibble)
library(dplyr)
```



# Quick start

```{r quickstart,eval=FALSE}
library(vtree2)

# example data set - same as Titanic, but with NAs
data(titanicNA)

# variables to show: Class and Survived
vt <- vtree(titanicNA, Class, Survived)

plot(vt, legend = TRUE)
plot(vt, layout = "proportional")
```

```{r quickstart_real,fig.width=10,fig.height=5,echo=FALSE,warning=FALSE,message=FALSE}
library(vtree2)

# example data set - same as Titanic, but with NAs
data(titanicNA)

# variables to show: Class and Survived
vt <- vtree(titanicNA, Class, Survived)

p1 <- plot(vt, legend = TRUE)
p2 <- plot(vt, layout = "proportional")
plot_grid(p1, p2)
```

For more information what the vtrees are and how to use them, see the [What
are Vtrees?](what_are_vtrees.html) vignette. Read on for the full user
manual.

# Vtree2 Manual

## Building vtree objects

### Case data

Vtrees are primarly built from *case data*: matrices or data frames in
which each row corresponds to a single sample. Each column is a data
variable and can be selected to be used and displayed on a vtree. For
example, consider the built-in data set `titanicNA`. This is the same data
set as `Titanic` from the `datasets` package, except that it is a cases
data frame and some of the values were replaced by NAs to simulate missing
data:


```{r vtree1}
data(titanicNA)
head(titanicNA)
```

As you can see, we have four variables: `Class`, `Sex`, `Age` and
`Survived`. Each of the `r nrow(titanicNA)` rows corresponds to a single
passenger.

The `vtree()` function then collects the levels of each selected variable
and counts the number of samples for each combination of levels:


```{r vtree2}
vt <- vtree(titanicNA, Class, Survived)
vt
```

Note that you select the variables that you want to consider in your vtree
- in the above example, we do not care about Sex or Age. Also note that the
variables follow the tidyverse logic - they are *data vars*, and do not
need to be quoted. If you need to use a variable name that is stored in
another variable, or if the variable names are not valid R identifiers, you
can use `tidyselect::all_of()` to select the variables.


```{r vtree3}
vars <- c("Class", "Survived")
vt <- vtree(titanicNA, all_of(vars))
```

That is one way to construct vtrees. The other is to use frequency tables.

### Frequency data

Often the data is already summarized in a frequency table, where each row
corresponds to a combination of levels of the selected variables, and there
is a column with the number of samples in which this combination occurs.

There are two functions to construct vtrees from frequency tables:
`vtree_from_freqtable()` and `cases_from_freqtable()`. The first one
constructs the vtree directly from the frequency table, while the second
one first expands the frequency table into a cases data frame which you can
use to construct the vtree with `vtree()`.

The original Titanic data set is a cross-tabulation of the four variables.
When converted to a data frame, it becomes a frequency table, with the
column `Freq` containing the number of passengers for each combination of
levels of the four variables:


```{r vtree4}
head(as.data.frame(Titanic))
cases <- cases_from_freqtable(Titanic)
vt <- vtree(cases, Class, Survived)

# or do it directly:
vt <- vtree_from_freqtable(Titanic, Class, Survived)
```

### Valid percentages

One of the important concepts to know about are the *valid percentages*. In
short, a valid percentage is the number of samples with a given value of a
variable divided by the total number of samples *for which this variable is
not missing*.

If you have 50 males, 50 females and 100 samples for which
the sex is unknown, then the number of males divided by the total number of
samples may be misleading (25%); valid percentages are calculated only from
the 100 samples for which the sex is known and the proportion of males and
females is 50% each.

The `vtree()` and `vtree_from_freqtable()` functions by default calculate
valid percentages; if you want to calculate the absoluet frequencies, use
the `.vp=FALSE`argument to these functions.

Once the vtree has been constructed, this cannot change.
Any operations downstream (like selecting and pruning nodes and modifying
their labels) will not change the calculated percentages.

### Vtree objects

You can inspect the vtree object with a number of methods.

```{r vtree5}
# names of the column variables associated with the nodes
names(vt)

# available levels of the different column variables:
levels(vt)

# summary frequency data for each node variable
summary(vt)

# check whether the vtree is calculated using valid
# percentages or absolute frequencies
is_vp(vt)
```

The vtree objects contain a data frame of nodes with a number of associated
columns, which include:

 * `node_key` - a unique identifier for each node
 * `node_id` - a unique integer identifier for each node
 * `node_col` - the name of the column variable associated with the node
 * `node_val` - the value of the variable associated with the node
 * `freq` - the percentage of samples within the node that have the given
   `node_val`
 * `n` - the number of samples in the current node
 * `missing` - the number of missing samples in the current node
 * `tot_n` - the number of samples in the parent node
 * `denom` - the number of samples within the node that have a non-missing value for the
   variable associated with the node. This is equal to `tot_n` only if the tree
   was not constructed with valid percentages. Otherwise, it is equal to
   the number of samples in the parent node that have a non-missing value
   for variable `node_col`.

Some other columns may be added later with `mutate()` or with the various `add_*()` functions, for example:

 * `label` - the label to be displayed on the node
 * `fill` - the fill color of the node
 * `color` - the text color of the label
 * `x`, `y`, `width`, `height` - layout information for the node

You can inspect the columns of the vtree object with `nodecols()`, which is
just a wrapper around `colnames(as_tibble(vt))`:


```{r vtree6}
nodecols(vt)
```

You can convert the vtree object to a data frame with `as_tibble`:

```{r vtree7}
as_tibble(vt)
```

### Vtree objects as graphs

The `vtree` class is a wrapper around the `tbl_graph` class from package
`tidygraph`. It is a directed graph, where each node corresponds to a
combination of levels of the selected variables. The `tbl_graph` class
itself inherits from `igraph`, which means that you can use all of the
`igraph` functions on `vtree` objects:


```{r vtree8}
igraph::degree(vt)
```

Since the default plotting function for `igraph` objects is replaced by
`vtree2`'s own `plot.vtree()`, if you want to use `igraph`'s plotting
function, you need to convert the vtree object to a `tbl_graph` object first:


```{r vtree9,fig.width=4,fig.height=4}
vt_tbl <- as_tbl_graph(vt)
plot(vt_tbl)
```


## Pruning, retain and selecting

### Pruning and keeping

Quite often, we don't want to show the whole tree, but only selected nodes.
For this, you can use the functions `prune()` and `retain()`. Each has a
few pecularities and defaults.

**Pruning** a node means: remove the node from the tree, along all the
nodes that follow it. However, with the parameter `follow_only=TRUE`, all
the following nodes will be pruned, but the given node will be kept.

**Retaining** means that you indicate which nodes you would like to keep,
and prune all other nodes.

Both pruning and retaining take a logical expression as an argument. The
expression is evaluated in a particular context: you can use both all the
node data frame columns (i.e., data vars such as `freq`, `n` or `path`), as
well as the node variables (like `Class` or `Survived`). All this works
naturally: you can specify that you want only to see the node where 
`Class == "1st"` or all nodes where `freq > 0.5`.

```{r nodes1,fig.width=9,fig.height=9}
# all variables except for Age
vt <- vtree_from_freqtable(Titanic, -Age)
p1 <- plot(vt)
# show only 1st class passengers
p2 <- vt |>
  retain(Class == "1st") |>
  plot(legend_tiny = FALSE, lwidth =.8)
# remove nodes with low frequency or number
p3 <- vt |>
  prune(freq < .2 | n < 20) |>
  plot(legend_tiny = FALSE, lwidth =.8)
# show only combinations of factors where more than a quarter of passengers
# survived
p4 <- vt |>
  retain(Survived == "Yes" & freq > .25) |>
  plot(legend_tiny = FALSE, lwidth =.8)
plot_grid(p1, p2, p3, p4, nrow = 2,
          labels=c("Original", "Retained Class:1st",
         "Pruned freq < .2 | n < 20",
         "Retained freq > .25 & Survived:Yes"),
         label_size = 10)
```


**Valid percentages and NA nodes.** There is one pecularity about how the
NA nodes are shown. If you try to prune a node corresponding to a missing
value, it will still be shown - if the tree was constructed with the
so-called "valid percentages" (`.vp=TRUE`, default). VP means that the
denominator used to calculate frequencies on the tree was the number of
*valid* samples, i.e. excluding the missing values. Therefore, without the
NA node shown for a VP-based tree, you lack information necessary to
understand the percentages shown.

```{r nodes2,fig.width=9,fig.height=6}
# all variables except for Age
vt <- vtree(titanicNA, -Age)
p1 <- vt |>
  # this should prune the NA nodes, but doesn't
  retain(Class %in% c("1st", "2nd")) |>
  prune(is.na(Sex)) |>
  plot(legend_tiny = FALSE, lwidth=.8)
vt_novp <- vtree(titanicNA, -Age, .vp = FALSE)
p2 <- vt_novp |>
  # this one prunes the NA nodes, b/c the tree 
  # is created with .vp = FALSE
  retain(Class %in% c("1st", "2nd")) |>
  prune(is.na(Sex)) |>
  plot(legend_tiny = FALSE, lwidth=.8)
plot_grid(p1, p2, nrow = 1,
          labels = c("VP",
                     "No VP"))
```

**Targetting nodes.** Quite often, you want to pick one particular node and
either prune it or select only that node for plotting.
The `path` column contains a human-readable specification of the node which
makes picking a certain node easy. For example, if you want to keep only
the node corresponding to males from 3rd class (and the following nodes),
you can ask for nodes where path is equal to `"Class:3rd/Sex:Male"`:

```{r nodes3,fig.width=8,fig.height=4}
vt <- vtree_from_freqtable(Titanic)
p1 <- plot(vt, legend_tiny = FALSE, dir="tb", show_root = FALSE)
p2 <- vt |>
  retain(path == "Class:3rd/Sex:Male") |>
  plot(legend = TRUE, dir="tb", show_root = FALSE)
plot_grid(p1, p2, nrow = 1, labels = c("Original tree", "Retained node"))
```

Note that the legend shown on the left in the above plot still contains all
the other variable levels, even if you can't see them on the tree. This is
on purpose: the stat summaries in the legend correspond to the whole data
set on which the tree was built, and not only the portion that is shown in
the tree.

### Finding and marking

## Creating summaries

## Creating layouts

Layouts will be automatically created by `plot()` if they are missing. The
parameters passed to the plot give some basic control over the layout:
`dir` determines the direction of the tree, `show_root` determines whether the
root node is shown, and `lwidth`/`lheight` determine the width and height
of the nodes relative to the available space.


```{r layouts1,fig.width=10,fig.height=5}
vt <- vtree_from_freqtable(Titanic, Class, Survived)
p1 <- plot(vt)
p2 <- plot(vt, dir = "tb", show_root = FALSE, lwidth = 0.8, lheight = 0.3)
plot_grid(p1, p2, nrow = 1)
```

The `plot()` function simply calls `add_layout()` to add the layout to the
vtree object with the specified parameters. Alternatively, you can call the
function `add_layout()` directly on the vtree object. The resulting vtree
contains the additional columns `x`, `y`, `width` and `height` for the
nodes and `x1`, `y1`, `x2`, `y2` for the edges. You can then modify these
columns directly.

In the following example we will modify a layout with pruned nodes. We will
move the nodes on the right (leafs) hand up a bit upwards.


```{r layouts3,fig.width=5,fig.height=5}
vt <- vtree_from_freqtable(Titanic, Class, Survived) |>
  prune(Class %in% c("1st", "2nd"), follow_only = TRUE) |>
  add_layout() |>
  mutate(y = ifelse(leaf, y + height/2, y))

# we need to modify the edges as well, but for that we need information
# from the nodes
nodes <- as_tibble(vt)

vt <- mutate(vt, x1 = nodes$x[from], y1 = nodes$y[from],
                 x2 = nodes$x[to] - nodes$width[to]/2, y2 = nodes$y[to],
             .edges = TRUE)

plot(vt)
```

Note how we mutate the edges with the `.edge = TRUE` argument. This is
different from `tidygraph`, where the object (tree) can either be in
activated "node" or "edge" mode. In `vtree2`, the object is always in
"node" mode (because it is more practical), but on the rare occassion that
you need to access the edges, you can.

The `add_layout()` function has additional parameters: `varspace` and
`varsize`. These can assign the amount of relative space available to the
node depending on the variable it is associated with. For example, we might
want to show a long summary which takes a lot of space for the leafs, but
only short labels for the internal nodes. You can tune the relative node
sizes with these two parameters.

```{r layouts4,fig.width=8,fig.height=2}
cases <- cases_from_freqtable(Titanic)
sm <- summary_vt(cases, vt, Age)

vt <- vtree(cases, Class, Sex, Survived) |>
  prune(Class == "Crew") |>
  add_labels() |>
  mutate(label = ifelse(leaf, paste0(label, "\n", sm), label)) |>
  add_layout(dir="tb", lheight=.8,
             varspace=c(root=1, Class=1,Sex=1,Survived=4))

# legend_tiny: not even column names on the margin
plot(vt, legend_tiny = FALSE)
```

Note that both `varspace` and `varsize` describe the node sizes along the
axis of the plot. That is, for horizontal layouts, they control the width
of the nodes; for vertical layouts, they control the height of the nodes.


## Adding column and value aliases

If you prefer to see a different name for the variable or its values on the
plot, there are basically two approaches.

First, you can just recode or rename the variables in the data from which
you build the vtree. This has an advantage: you ensure that the labels
you see on the plot are the same as in the original data. For example, you
can do the following:


```{r aliases1,fig.width=4,fig.height=4}
cases <- as_tibble(Titanic) |>
  mutate(Class = recode(Class, "1st" = "First", "2nd" = "Second",
                        "3rd" = "Third")) |>
  cases_from_freqtable(.freq_col = "n")

vtree(cases, Class, Survived) |> plot()
```

However, some times what you want to see on the plot is not convenient to
work with in the data, for example if you want to show a longer description
of the possible values, add Unicode characters or markdown / HTML
formatting for rich text labels. In this case, you can specify an alias for
either using the `col_alias` and `val_alias` arguments of `add_aliases()`.
Not all variables or all levels (values) need to be specified, only those
you want to change:

```{r aliases2,fig.width=8,fig.height=4}
col_alias <- list(Class = "Passenger\nClass",
                  Sex = "Gender",
                  Age = "Age\nGroup",
                  Survived = "Survival\nStatus")

val_alias <- list(NAs = "Missing",
                  Class = c("1st" = "First",
                             "2nd" = "Second",
                             "3rd" = "Third"))

vtree(titanicNA, Class, Survived) |>
  add_aliases(col_alias = col_alias,
             val_alias = val_alias) |>
  plot(legend = TRUE, dir="tb", show_root = FALSE,
       margins = c(0.05, 0.05, 0.05, 0.10))
```

The special `NAs` name is used for all missing values, regardless of the variable.

Unnecessary gory details:

 * `add_aliases()` and `add_labels()` always ensure that columns
   `col_alias` and `val_alias` are present in the vtree object.
 * If no custom format is specified, these two columns will be identical to
   the `node_col` and `node_val` columns, respectively.
 * In addition, the alias information is stored in the `alias`
   attribute of the vtree object, which is subsequently used
   by `plot()` to create the legend.

## Adding and modifying labels

Labels to be plotted are taken from the `label` column of the vtree object.
If this column is missing, `plot()` calls `add_labels()` to add default
labels. For custom labels there are three routes:

 * directly add a label column to the vtree object with `mutate()`
 * use `add_labels()` with a custom format
 * first use `add_labels()`, then modify the label column with `mutate()`

### Directly creating a label column with `mutate()`

Since the `mutate()` function is called in the context of the vtree node
data frame, you can use any of the columns of the vtree object directly.
For example, you can show the `node_key` column, a unique identifier for
each node.

```{r labels1,fig.width=8,fig.height=4}
vtree(titanicNA, Class, Survived) |>
  mutate(label = paste0("node_key:\n", node_key)) |>
  plot(legend_tiny = FALSE)
```

### Using `add_labels()` to add default or custom labels

The `add_labels()` function can produce two types of default labels:
"simple" and "long", chosen through the `template` argument:

```{r labels2,fig.width=8,fig.height=4}
vt <- vtree(titanicNA, Class)
p1 <- add_labels(vt) |> plot() # template = "simple"
p2 <- add_labels(vt, template = "long") |> plot()
plot_grid(p1, p2, nrow = 1)
```

These labels are derived directly from columns of the vtree object:
`node_col`, `node_val`, `freq` and `n`.

### Using custom formatting

Formatting can also be done with the `fmt`/`fmt_na` parameters, which are
R expressions. You can use sprintf, glue, paste or whichever
expressions you like to construct a label from the following variables:

 * `freq`, the frequency for a node
 * `n`, number of samples of a node
 * `node_col`, name of the variable associated with a node
 * `node_val`, value of the variable associated with a node
 * `col_alias`,`val_alias`: if you have defined aliases with
   `add_aliases()`, you will find them here. Otherwise the columns are
   same as `node_col` and `node_val`.
 * plus whatever new columns you have added to the vtree with mutate().

You can list the columns in the node data frame of a vtree object with
`nodecols(vt)` (or simply `colnames(as_tibble(vt))`).

If provided, `fmt_na` is used to format the NA nodes.

Here is simple example using `glue()`:

```{r labels3,fig.width=4,fig.height=4}
library(glue)
vt <- vtree(titanicNA, Class, Sex) |>
  add_labels(fmt =
    glue("{node_col}: {node_val}\nfreq={format(freq, digits=2)}"),
             fmt_na =
    glue("{node_col}: Missing\nfreq={format(freq, digits=2)}"))
plot(vt, legend_tiny = FALSE)
```

Note that the root node also got a label, but since `node_col` and `node_val` are both "" for the root, the label is not very informative. We can change it:

```{r labels4,fig.width=8,fig.height=4}
library(glue)
vt <- vtree(titanicNA, Class, Sex) |>
  add_labels(fmt =
    glue("{node_col}: {node_val}\nfreq={format(freq, digits=2)}\nn={n}"),
             fmt_na =
    glue("{node_col}: Missing\nfreq={format(freq, digits=2)}\nn={n}")) |>
  mutate(label = ifelse(path == "root",
                        glue("All passengers\nn={n}"),
                        label))
plot(vt, legend_tiny = FALSE, dir="tb")
```

### Using a mask

You can specify a mask with `add_labels()` - a logical vector with the same
length as the number of nodes - to indicate which nodes should have their
labels modified. For example, you may want a different information for the
leaf nodes as for other nodes:

```{r labels_mask,fig.width=4,fig.height=4}
vt <- vtree_from_freqtable(Titanic, Class, Sex, Survived) |>
  retain(path == "Class:1st") # only keep the first class passengers
mask <- find_nodes(vt, leaf) # leaf is a logical vector
add_labels(vt, mask = mask, template = "long") |>
  add_labels(mask = !mask, fmt = glue("{node_val}")) |>
  plot(legend_tiny = FALSE, dir="tb", show_root = FALSE,
       lwidth = 0.5, lheight = 0.5)
```

## Adding and modifying colors

## Plotting

Objects of type vtree can be directly plotted with the `plot()` method. If
the labels, colors and layouts are missing, then plot adds some automatic
defaults by running `vtree |> add_labels() |> add_colors() |> add_layout()`
before plotting. Some of the arguments for these functions can be passed
directly to the `plot()` function.

### Layouts

Currently, there are three layouts implemented: "regular", "flushed_left",
"flushed_right" and "proportional". The default layout is "regular", which
simply shows the tree structure with all nodes having the same sizes. The
"flushed_*" layouts are similar, but the nodes are always flushed to one side
of the plot.

The proportional layout shows the nodes with sizes proportional to the
number of observations in that node.


```{r layouts2,fig.width=8,fig.height=8}
# the default column name for .freq_col is "Freq", same as in Titanic,
# so no need to specify it here
vt <- vtree_from_freqtable(Titanic, Class, Sex, Survived)
p1 <- plot(vt) # layout regular
p2 <- plot(vt, layout = "proportional")
p3 <- plot(vt, layout = "flushed_left", dir="tb")
p4 <- plot(vt, layout = "flushed_right", dir="tb")
plot_grid(p1, p2, p3, p4, nrow = 2)
```

Layouts are added with the `add_layout()` function, which you can call
directly on the vtree object, resulting in a new vtree object with
additional columns for the layout.

With the `layout_func` argument, you can specify a custom function to
calculate the layout yourself. For example, this function might call
`add_layout()` first and then adjust the calculated positions of the nodes
in some way. See the documentation for `add_layout()` for details.

### Colors, fill colors and labels

Colors can be added with the `add_palette()` function and then modified
by directly manipulating the vtree object with `mutate()`. Each node can
have the `fill` and `color` columns, corresponding to the fill color of the
node and the text color of the node label. 

If `fill` is missing from the vtree object, `plot()` will call
`add_palette()` to assign fill colors automatically. If `color` is missing, but
`fill` is present, then white or black will be chosen automatically
depending on the contrast with the fill color.

```{r colors1,fig.width=8,fig.height=4}
pal <- colorRampPalette(c("white", "steelblue"))(101)

p1 <- vt |>
  mutate(fill = pal[round(freq * 100) + 1]) |>
  plot()

p2 <- vt |>
  mutate(abs_freq = n / max(n)) |>
  mutate(fill = pal[round(abs_freq * 100) + 1]) |>
 plot()

cowplot::plot_grid(p1, p2)
```

Colors can be also assigned automatically based on names of RColorBrewer
palettes. There is a default order of palettes for the variables, but you
can override it with the `palettes` argument of `plot()` or
`add_palette()`.

Similarly, if the `label` column is missing from the vtree object, `plot()`
will call `add_labels()` (see above) to assign labels automatically.

### Legends with summaries

The legend shows a summary for each variable underneath or next to the
part of the tree which corresponds to this variable. Variable summaries are
calculated when vtree is constructed and stored in the `summary` attribute
of the vtree object. You can print them with `summary(vtree_object)`:

```{r summaries1}
vt <- vtree_from_freqtable(Titanic, Class, Sex, Survived)
summary(vt)
```

The summaries will not change if you prune the tree, since they are tied to
the data and not the tree structure:


```{r summaries2,fig.width=8,fig.height=4}
mar <- c(0.05, 0.05, 0.33, 0.05)
p1 <- plot(vt, legend = TRUE, lwidth=.8, margins = mar)
p2 <- retain(vt, path == "Class:1st") |>
  plot(legend = TRUE, lwidth=.8, margins = mar)
plot_grid(p1, p2, nrow = 1)
```


### Other `plot()` arguments

**`lwidth`, `lheight`** - label width and height relative to available space.
The layout functions calculate the available space for each node, and then
these parameters determine how much of that space is used for the label.

For the proportional layout, `lheight` is ignored, since the height of the
node is determined by the number of observations in that node.

**`margins`** - a numeric vector of length 4, specifying the top, right,
bottom and left margins as fractions of the total device space. So, for
example, for 10% margins on all sides, use `margins = c(0.1, 0.1, 0.1, 0.1)`.

**`show_root`** - if TRUE (default), show the root node (total population).

**`legend_tiny`** - if FALSE, not even minimal variable legend is shown on the
plot. If TRUE (default), the variable (column) names are shown on the margin.

```{r fig.width=5,fig.height=5}
vt <- vtree_from_freqtable(Titanic, Class, Sex, Survived)
plot(vt, legend_tiny = FALSE)
```

**`dir`** direction of the tree. One of "lr" (left to right), "rl"
(right to left), "tb" (top to bottom), "bt" (bottom to top).
Default is "lr".

```{r fig.width=20,fig.height=5}
p1 <- plot(vt) # same as dir = "lr"
p2 <- plot(vt, dir = "tb")
p3 <- plot(vt, dir = "bt")
p4 <- plot(vt, dir = "rl")
plot_grid(p1, p2, p3, p4, nrow = 1)
```

**`fontsizes`** Font sizes for the various labels are automatically fit to
the available space, but sometimes you might manually adjust them. The
`fontsizes` is a named list with the following optional fields: 
 * `nodes`: node labels (default: "fixed" for regular layout,
            "adaptive" for proportional layout)
 * `var_labels`: variable labels (default: "fixed")
 * `legend_labels`: for variable levels when `legend=TRUE` (default: "fixed")

Each element can be either a number, in which case the given font size is
directly used, or either "fixed" or "adaptive". When "fixed" is used, then
optimal font size is calculated for all objects within the group, and then
the minimal font size is used for all objects. When "adaptive" is used, then each
object is fit separately, and the font sizes can be different for each object.

**`lwd`** line width for use with plotting. Unfortunately, the default line
width is not relative to the device size, but is fixed to an absolute
value; therefore, for small devices the lines may appeary too thick.

### Rich text labels

It is possible to use simple HTML or markdown formatting in the labels and
then use `richtext=TRUE` in the `plot()` function to render the labels with
the formatting. The plot rendering is noticably slower, but it allows for
more flexibility in the label formatting.

```{r richtext1,fig.width=4,fig.height=4}
vt <- vtree_from_freqtable(Titanic, Class, Survived) |>
  add_labels() |>
  mark(path == "Class:1st/Survived:No") |>
  mutate(label = ifelse(mark, paste0('**', label, '**'), label)) |>
  mark(path == "Class:2nd/Survived:No") |>
  mutate(label = ifelse(mark,
                        paste0('<span style="color:red">', label, '</span>'),
                        label))

plot(vt, richtext = TRUE, legend = TRUE)
```

You can also include the rich text formatting in the aliases for the
variable names and values, which will then be used both in the labels and
in the legend.

```{r richtext2,fig.width=6,fig.height=4}
vt <- vtree_from_freqtable(Titanic, Class, Sex, Survived)

# we can use aliases through add_labels() to change the formatting of the
# variable names and values
cols <- list(Class = "**Class**", Sex = "**Sex**",
             Age = "**Age**", Survived = "**Lived**")
vals <- list(Class = c("1st" = "1<sup>st</sup>", "2nd" = "2<sup>nd</sup>",
                  "3rd" = "3<sup>rd</sup>"),
            Sex = c("Male" = "**<span style='color:blue'>Male</span>**",
                    "Female" = "**<span style='color:lightblue'>Female</span>**"))
vt <- add_aliases(vt, col_alias = cols, val_alias = vals)

plot(vt, richtext = TRUE, legend = TRUE)
```





## Inset plots

It is possible to add any kind of "grob" - graphical object - to the nodes
description. This includes not only bitmap images, but also `ggplot2`
plots, which can be easily converted to a grob with the function
`ggplot2::ggplotGrob()`.

The important thing to remember is that the inset graphics (1) must be a
grob and (2) must be stored in a list-column of the vtree object called
"grob".

In the following example, we will download three images from Wikipedia
corresponding to the three species of iris flowers from the famous Fisher
data set. We will show on the vtree how many of each of the species have
long petals, and we will illustrate the Species nodes with the
corresponding images.

```{r inset1,fig.width=7,fig.height=4}
library(dplyr)
library(purrr)

iris_urls <- list(
versicolor="https://upload.wikimedia.org/wikipedia/commons/thumb/2/27/Blue_Flag%2C_Ottawa.jpg/500px-Blue_Flag%2C_Ottawa.jpg",
setosa="https://upload.wikimedia.org/wikipedia/commons/thumb/a/a7/Irissetosa1.jpg/500px-Irissetosa1.jpg",
virginica="https://upload.wikimedia.org/wikipedia/commons/thumb/f/f8/Iris_virginica_2.jpg/500px-Iris_virginica_2.jpg")

get_grob <- function(url) {
  tf <- tempfile(fileext = ".jpg")
  download.file(url, tf, mode="wb")
  img <- jpeg::readJPEG(tf)
  grid::rasterGrob(img)
}

iris_grobs <- lapply(iris_urls, get_grob)

vt <- iris |>
  mutate(Long_Petals = as.character(Petal.Length > 4)) |>
  vtree(Species, Long_Petals) |>
  add_labels() |>
  mutate(label = ifelse(leaf, label, paste0("Iris\n", node_val)))
vt <- vt |>
  mutate(grob = map(pull(vt, "node_val"), ~ iris_grobs[[.x]]))

layout <- vt |>
  add_palette(palettes = c("Greens", "Blues")) |>
  add_layout(dir="tb", show_root=FALSE, lwidth=.9, lheight=.8,
             varspace=c(Species=4,Long_Petals=1))
layout |>
  plot(margins=c(.05, .05, .05, .2))
```

Here, we choose only the "Sex" and "Survived" nodes for the construction of
the vtree. However, we make a ggplot object for each of the leaf nodes
("Survived:Yes" and "Survived:No") and insert the resulting graphics object
into the node.

For this, we create the `grob` column and convert the ggplot2 to a grob
with `ggplot2::ggplotGrob()`. The `grob` column is a list-column, so we need to
manipulate it carefully. Also, we need to set an NA value for the non-leaf
nodes.

```{r insets2,fig.width=10,fig.height=4}
# first, a tree with simplified labels on the leaf nodes
cases <- cases_from_freqtable(Titanic)
vt <- vtree(cases, Sex, Survived)

library(ggplot2)

# minimalistic ggplot showing how many passengers were in each class
# of the data frame df.
clasplot <- function(df) {
  ggplot(df, aes(x = Class, fill = Class)) +
    geom_bar() +
    scale_fill_discrete(drop=FALSE) +
    scale_x_discrete(drop=FALSE) +
    ylim(0, max(table(cases$Class))) +
    theme_minimal() +
    # remove labels, legend and axes
    theme(legend.position = "none",
          axis.title = element_blank(),
          axis.ticks = element_blank())
}

# use vtree_apply to produce plots
plots <- vtree_apply(cases, vt, FUN = clasplot)
grobs <- lapply(plots, ggplot2::ggplotGrob)
grobs[!pull(vt, leaf)] <- NA # plots only for the leaf nodes

# we add layout manually, so we can control it better
# this varspace arg means: reserve 4x as much space for the Survived nodes
# as for the Sex nodes
vt <- vt |> 
  add_layout(dir="tb", show_root=FALSE, lheight=.8,
             varspace=c(Sex=1,Survived=3)) |>
  mutate(shape = ifelse(leaf, "rectangle", "roundrectangle")) |>
  mutate(grob = grobs)
plot(vt)
```

There is one issue with such plots: unlike `vtree2`, `ggplot2` does not fit
font sizes automatically to the available space. If your image is too
small, bad things will happen:


```{r fig.width=5,fig.height=2.5}
plot(vt)
```



