---
title: "What are vtrees?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What are vtrees?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup,echo=FALSE,warning=FALSE,message=FALSE}
library(vtree2)
library(cowplot)
library(tibble)
library(dplyr)
```


# Vtrees as representations of categorical data

The main purpose of vtrees is to understand and visualize hierarchical,
categorical data. A vtree is a tree diagram that shows how the data is split
by the levels of categorical variables.

In the example most frequently used in the vtree package, the Titanic data
set, we have several categorical variables, such as `Class` (passenger
class), `Sex` (gender) and `Survived` (whether a passenger survived the
sinking of the Titanic). You may ask, for example, how many female
passengers who were traveling in the first class survived? What was 
their percentage among all first class female passengers?

```{r repr1}
cases <- cases_from_freqtable(Titanic)
head(cases)

# female 1st class survivors
sum(cases$Class == "1st" & cases$Sex == "Female" & cases$Survived == "Yes")

# percentage
sum(cases$Class == "1st" & cases$Sex == "Female" & cases$Survived == "Yes") /
  sum(cases$Class == "1st" & cases$Sex == "Female") * 100
```

Rather than calculating these numbers manually, we can visualize all that
in one go on a vtree:

```{r repr2,fig.width=6,fig.height=5,warning=FALSE,message=FALSE}
vt <- vtree(cases, Class, Sex, Survived)

plot(vt)
```

If you follow the path from the root node on the left, to the first class,
to the female passengers, and then to the survivors, you will notice the
percentage of survivors among the first class is what we calculated above:

```{r repr3,fig.width=6,fig.height=5}
vt |>
  mark(path %in% c("root", "Class:1st", "Class:1st/Sex:Female",
                   "Class:1st/Sex:Female/Survived:Yes")) |>
  mutate(fill = ifelse(mark, "red", "white")) |>
  plot()
```

In a way, this is reminding of a consort diagram. We can style the above
plot so it looks a bit more like a consort diagram:

```{r repr4,fig.width=8,fig.height=5}
vt |>
  mutate(fill = "white") |>
  plot(dir="tb", layout="flushed_right", legend_tiny = FALSE)
```

# FP / NPV example

The following example may convince you of the use vtree representation can
have for your data.

```{r fpnpv_prep,echo=FALSE}
FPR <- .02 # p that healthy is positive
FNR <- .02 # p that infected is negative
prevalence <- 1/50 
N <- 50000

data <- tribble(
 ~ Status, ~ Test, ~ N,
 "Infected", "Positive", round(prevalence * N * (1-FNR)),
 "Infected", "Negative", round(prevalence * N * FNR),
 "Healthy",  "Positive", round((1 - prevalence) * N * FNR), 
 "Healthy",  "Negative", round((1 - prevalence) * N * (1 - FNR))
 )
```

Imagine a disease and a test that we have for that disease. How good is the
test?

We can consider how often the test makes an error for an infected subject,
incorrectly giving the negative result. This is called a false negative
(FN). The probability of this error is called false negative rate (FNR),
and its compliment the sensitivity: the more sensitive our test is, the
less likely that a test result will be negative if the person is sick.
Let's assume that our test is highly sensitive, for example that the FNR
is `r format(100 * FNR)`% (making sensitivity `r format(100 * (1-FNR))`%).


Another type of the error is if we have a healthy subject, but the test is
showing as positive. This value is the number of false positives (FP)
divided by the total number of healthy people, also known as FPR - false
positive rate. The complement of this value ($1 - FPR$) is called
specificity. Say, we have an outstanding test and the specificity is 
`r format(100 * (1-FPR))`%, so the false positive rate is
`r format(100 * FPR)`%.

OK, but there is more to the story. Whether a positive test result
corresponds to a real patient or a false positive depends on how likely is
that the person is actually healthy in reality. This is disease prevalence,
how often the disease is encountered when people are tested. Say, that the
disease is quite common, and the prevalence is
`r format(100 * prevalence)`%.

From this, assuming we performed screening tests on `r N` persons in a
population, we get the following table:


```{r npv_table,echo=FALSE}
data |> knitr::kable()
```

Here is how we can build our data:

```{r fpnpv_prep_noeval,eval=FALSE}
FPR <- .02 # p that healthy is positive
FNR <- .02 # p that infected is negative
prevalence <- 1/50 
N <- 50000

data <- tribble(
 ~ Status, ~ Test, ~ N,
 "Infected", "Positive", round(prevalence * N * (1-FNR)),
 "Infected", "Negative", round(prevalence * N * FNR),
 "Healthy",  "Positive", round((1 - prevalence) * N * FNR), 
 "Healthy",  "Negative", round((1 - prevalence) * N * (1 - FNR))
 )
```


We can visualize this table as a vtree. Since it is a frequency table, we
need to use the `vtree_from_freqtable()` function[^1]. For starters, we will
show the Status first, and then the Test result. This will show us how many
of the healthy persons were incorrectly classified as positive, and how
many of the infected persons were incorrectly classified as negative.

[^1]: `vtree()` works with cases data frames, where each row is a single
  observation. Here we have a frequency table, in which each row is a group
  of observations with the same levels of variables.


```{r fpnpv1,fig.width=10,fig.height=4}
vt <- vtree_from_freqtable(data,
                           Status, Test,
                           .freq_col = "N")
plot(vt)
```

As you can see, the percentages on the leaf nodes (right side of the plot)
correspond to our FPR and FNR values. The percentages on the internal nodes
correspond to the disease prevalence.

What if we were to inverse this question? What is the fraction of healthy
people among those who tested positive, and vice versa - what is the
fraction of infected people among those who tested negative?


```{r fpnpv2,fig.width=10,fig.height=4}
vt2 <- vtree_from_freqtable(data,
                           Test, Status,
                           .freq_col = "N")
plot(vt2)
```


```{r echo=FALSE}
npv <- as_tibble(vt2) |> filter(path == "Test:Negative/Status:Healthy") |>
  pull(freq)
ppv <- as_tibble(vt2) |> filter(path == "Test:Positive/Status:Infected") |>
  pull(freq)
```


We see that among those who were tested negative, the large majority
(practically `r round(100 * npv)`%) were indeed healthy. This is the
negative predictive value (NPV) of the test.

However, among those who were tested positive, only `r format(100 * ppv)`%
were actually infected. This is the positive predictive value (PPV) of the
test and it shows that despite the test being very specific and sensitive,
when a person has a positive test result, there is more than a
`r format(100 * (1 - ppv))`% chance that the person is actually healthy.


## Vtree workflow

In `vtree2`, the workflow is split into several steps:

 * Prepare the data (outside of `vtree2`)
 * Build the vtree object with `vtree()` or `vtree_from_freqtable()`
 * (Optional) Prune, retain or select nodes with `prune()`, `retain()` and
   friends.
 * (Optional) Create summaries with `summary_vt()`.
 * (Optional) Add aliases for variable names and values with `add_aliases()`.
 * (Optional) Add or modify labels with `add_labels()` and `mutate()`.
 * (Optional) Add or modify colors with `add_palette()` and `mutate()`.
 * Plot the vtree with `plot()`.

Here is an example demonstrating all these steps based on the `titanicNA`
data set. This data set is the same as the Titanic data set, but with some
missing values randomly added for demonstration purposes.


```{r workflow_example1,fig.width=10,fig.height=5}
data(titanicNA)

vtree(titanicNA, Class, Sex, Survived) |>
  # only retain the third class passengers
  retain(path == "Class:3rd") |>
  # change how variables are displayed
  add_aliases(val_alias = list(Class = c("1st" = "First",
                                    "2nd" = "Second",
                                    "3rd" = "Third"))) |>
  # add default labels, built from aliases
  add_labels() |>
  # change the labels for the missing values
  mutate(label = gsub("NA", "Missing", label)) |>
  # add colors
  add_palette(palettes = c("Greys", "Blues", "Purples"),
              na_fill = "grey90") |>
  # change the color for Females who survived
  # mark() is a helper function that returns TRUE for the nodes that
  # match the given condition
  mark(path == "Class:3rd/Sex:Female/Survived:Yes") |>
  mutate(fill = ifelse(mark, "red", fill)) |>
  mutate(color = ifelse(mark, "white", color)) |>
  # plotting with legend and custom margins
  plot(legend = TRUE,
       margins = c(0.05, 0.05, 0.25, 0.05))
```





